---
title: "lexsync: a cross-platform pipeline for multidimensional lexical optimisation and hardware-timed experiment generation"
author:
  - name: Pablo Bernabeu
date: today
abstract: |
  Psycholinguistic stimulus preparation is fragmented across tools that match
  words, tools that present them and the separate R and Python ecosystems in
  which those tools live. We present 'lexsync', a research-grade toolkit,
  released as a 'CRAN'-ready R package and a structurally identical
  'PyPI'-ready Python package, that unifies the whole path from corpus to
  laboratory. From a word-frequency corpus in any of dozens of languages
  'lexsync' selects stimuli that are matched in parallel across several lexical
  dimensions (length, frequency, orthographic neighbourhood density and OLD20),
  counterbalances them, reports the realised match quality and then generates
  ready-to-run 'PsychoPy' and 'OpenSesame' experiment scripts with hardware
  triggers injected at stimulus onset for electroencephalography. The matching
  is deterministic, so the two engines return identical stimuli; we demonstrate
  this on four worked examples in English and Spanish. The generation step
  requires no presentation software and no hardware, which keeps the whole
  workflow reproducible. 'lexsync' generalises the FAIR artificial-grammar
  workflow of an existing longitudinal study and is distributed under open
  licences with a machine-readable corpus registry.
bibliography: references.bib
format:
  html:
    toc: true
    number-sections: true
    df-print: kable
    embed-resources: true
  docx: default
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
have_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
find_out <- function(p) {
  for (base in c("../output", "output")) {
    f <- file.path(base, p)
    if (file.exists(f)) return(f)
  }
  stop(sprintf("lexsync manuscript: output file not found: %s", p))
}
read_out <- function(p) read.csv(find_out(p), stringsAsFactors = FALSE, fileEncoding = "UTF-8")

designs <- c("en_freqcontrast_english", "en_ndensity_english",
             "es_freqcontrast_spanish", "es_ndensity_spanish")
labels <- c(
  en_freqcontrast_english = "English: frequency contrast",
  en_ndensity_english     = "English: neighbourhood contrast",
  es_freqcontrast_spanish = "Spanish: frequency contrast",
  es_ndensity_spanish     = "Spanish: neighbourhood contrast"
)
comps <- do.call(rbind, lapply(designs, function(d) {
  x <- read_out(file.path("reports", paste0(d, "_comparisons_R.csv")))
  x$design <- labels[[d]]
  x
}))

# Readable dimension labels and the role of each dimension in each design.
dim_labels <- c(frequency = "Frequency (Zipf)", length = "Length (letters)",
                n_density = "Neighbourhood N", old20 = "OLD20")
find_cfg <- function(f) {
  for (b in c("../config", "config")) {
    p <- file.path(b, f); if (file.exists(p)) return(p)
  }
  stop(f)
}
cfg_files <- c(
  "English: frequency contrast"     = "design_en_freqcontrast.yaml",
  "English: neighbourhood contrast" = "design_en_ndensity.yaml",
  "Spanish: frequency contrast"     = "design_es_freqcontrast.yaml",
  "Spanish: neighbourhood contrast" = "design_es_ndensity.yaml"
)
controlled_map <- lapply(cfg_files, function(f) unlist(yaml::read_yaml(find_cfg(f))$match_on))
names(controlled_map) <- names(cfg_files)

comps$absd <- abs(comps$cohens_d)
comps <- do.call(rbind, lapply(split(comps, comps$design), function(x) {
  ctrl <- controlled_map[[x$design[1]]]
  x$role <- ifelse(x$dimension %in% ctrl, "Controlled", "Free (co-varying)")
  free <- which(x$role != "Controlled")
  if (length(free)) x$role[free[which.max(x$absd[free])]] <- "Manipulated"
  x
}))
comps$dim_label <- factor(dim_labels[comps$dimension],
                          levels = c("Frequency (Zipf)", "Length (letters)", "Neighbourhood N", "OLD20"))
comps$role <- factor(comps$role, levels = c("Manipulated", "Controlled", "Free (co-varying)"))

pretty_desc <- function(d) {
  x <- read_out(file.path("reports", paste0(d, "_descriptives_R.csv")))
  x$group <- gsub("_", " ", x$group)
  x$dimension <- as.character(dim_labels[x$dimension])
  names(x) <- c("Condition", "Dimension", "n", "Mean", "SD", "Min", "Median", "Max")
  x
}
```

# Introduction

The credibility of psycholinguistic and neurolinguistic research rests on
carefully controlled stimuli and on precise timing. Both are laborious to
achieve, and the tools that support them are scattered. A researcher who wishes
to contrast, say, high- and low-frequency words while equating length and
orthographic neighbourhood typically matches the items in one program, lays out
the trial structure in another and writes the trigger code that aligns each word
with the electroencephalography (EEG) recording by hand. Each hand-off is an
opportunity for error and a barrier to reuse. The 'FAIR' guiding principles
[@wilkinson2016fair] — that research objects be findable, accessible,
interoperable and reusable — apply as much to this preparatory machinery as to
the data it ultimately yields, yet stimulus pipelines are seldom shared in a
form that others can run end to end.

Two families of tool address parts of the problem. The first matches or equates
word stimuli on lexical properties. 'LexOPS' [@taylor2020lexops] generates
controlled word lists for arbitrary factorial designs, with both per-dimension
tolerances and weighted multivariate Euclidean matching; 'LIBRA'
[@lintz2021libra] equates lists across conditions using a resampling algorithm;
and 'LASTU' [@itkonen2024lastu] provides searchable access to a language-specific
lexicon. These tools are mature and widely used, but they end at the stimulus
list: none generates the experiment that presents it, and most are tied to a
single language and to one programming ecosystem. The second family builds and
runs experiments. 'PsychoPy' [@peirce2019psychopy2] and 'OpenSesame'
[@mathot2012opensesame] offer precise spatial and temporal control over stimulus
presentation and support the hardware triggers that ERP research requires, but
they take the stimulus set as given and do nothing to construct or to balance it.

The gap, then, is not a missing capability but a missing integration. No single,
installable framework spans many-language corpus access, parallel
multidimensional matching, counterbalancing and the automated generation of
hardware-timed presentation scripts, and none does so across both the R and the
Python ecosystems and both major open presentation platforms. A second, subtler
gap concerns timing. In the artificial-grammar workflow that motivated this work
[@gonzalezalonso2025l3], EEG markers were sent from an experiment item that
followed the word, so that only the first marker was even approximately aligned
with stimulus onset; the others were sequence-ordered. This is a common pattern
and an avoidable source of jitter.

We present 'lexsync', a toolkit that fills these gaps. It is released as two
independent but structurally identical packages — a 'CRAN'-ready R package and a
'PyPI'-ready Python package — that share module names, function names and
behaviour. From a corpus 'lexsync' selects matched stimuli, counterbalances and
documents them, then writes a 'PsychoPy' script in which each onset trigger is
bound to the exact screen flip on which the word appears, using
`win.callOnFlip`, together with a complete 'OpenSesame' experiment. Crucially the
matching is deterministic, so the R and Python engines return identical stimuli.
'lexsync' generalises the materials workflow of @gonzalezalonso2025l3 into a
reusable, FAIR-compliant tool.

# The lexsync packages

## Architecture and cross-language design

Each package is organised into the same modules: `querying` (load a lexicon and
compute lexical dimensions), `matching` (the constraint engine), `counterbalancing`,
`validation` (match-quality reporting), `scripting` (experiment generation),
`logging` (the automated run log) and `corpora` (the registry). A single
orchestrator, `run_pipeline`, drives them in an explicit loop over designs and
languages, so that adding a language or a design is a matter of supplying a
configuration file and a lexicon rather than writing code. The two packages are
deliberately lean: the R package depends only on 'readr', 'yaml', 'stringdist'
and 'jsonlite' beyond base R, and the Python package on 'pandas', 'numpy',
'PyYAML', 'rapidfuzz' and 'scipy'.

## A many-language corpus registry

Stimuli are only as good as the corpus behind them. 'lexsync' reads a
machine-readable registry that records, for each supported corpus, its language,
access URL, licence and citation. Two connectors are provided. The first exposes
the curated SUBTLEX family and the 'openlexicon' collection — for example
SUBTLEX-UK for British English [@vanheuven2014subtlexuk], SUBTLEX-ESP for Spanish
[@cuetos2011subtlexesp] and SUBTLEX-US [@brysbaert2009subtlexus] — each
individually citable and redistributable under CC BY-SA terms. The second is the
'wordfreq' package [@speer2022wordfreq], which yields frequencies for roughly
forty languages through a single dependency. The demonstrations below build their
English and Spanish lexica with 'wordfreq', which aggregates subtitle and web
corpora including the SUBTLEX lists; the registry documents the curated corpora as
alternatives. Small example lexica are bundled inside each package for offline use
and testing, while full corpora are fetched on demand into a user cache, keeping
the installed packages within distribution size limits.

## The multidimensional matching algorithm

For every word 'lexsync' records its length, its frequency and two
orthographic-neighbourhood measures: Coltheart's N, the number of words of the
same length differing by a single letter [@coltheart1977access], and OLD20, the
mean Levenshtein distance to the twenty nearest neighbours [@yarkoni2008old20].
The neighbourhood measures are computed transparently from the lexicon rather
than taken on trust.

Matching proceeds in two deterministic stages. First, a tolerance pre-filter
keeps only candidates that fall within `mean ± k·SD` of the anchor condition on
each control dimension, generalising the windowed selection of the original
workflow. Second, every remaining condition is matched to the anchor item by
item: each dimension is standardised, the standardised Euclidean distance from
each candidate to the target item is computed and the nearest unused candidate is
assigned, with ties broken by a stable, locale-independent key. Because no random
number generator is used and string comparisons are performed in byte order in
both engines, the R and Python packages select identical stimuli from identical
input — a parity we verify below.

## Counterbalancing and reporting

Matched items are assigned to presentation lists and a reproducible trial order,
and a participant table crosses any counterbalancing factors. 'lexsync' then
writes a match-quality report — per-condition descriptive statistics for every
dimension, standardised mean differences against the anchor and a two one-sided
tests (TOST) equivalence test [@lakens2017equivalence] — and an automated,
comprehensive run log recording each stage, its parameters, the seed, package and
corpus provenance and a fingerprint of every file written. The report makes the
realised control explicit, in the spirit of the original workflow's balance
checks, and the run log makes a run auditable.

## Hardware time-locking

The 'scripting' module emits two presentation targets from the same matched
stimuli. The 'PsychoPy' script binds each onset trigger to the stimulus flip:

```python
win.callOnFlip(port.setData, target_trigger)   # written on the onset flip
for frame in range(WORD_DURATION_FRAMES):
    word_stim.draw()
    win.flip()
    if frame == RESET_AFTER_FRAMES:
        win.callOnFlip(port.setData, 0)
```

A mock port prints the codes when no parallel-port driver is present, so the
script runs end to end with no hardware. The 'OpenSesame' export is a complete,
plain-text `.osexp`, generated programmatically so that its tab-sensitive format
is always valid and modelled on a proven-working experiment from
@gonzalezalonso2025l3; it sends the marker from an inline script placed
immediately after the word, supporting both a parallel-port and a serial backend
with a test-mode fallback. The two backends differ in an instructive way: the
'PsychoPy' path is genuinely flip-locked, whereas 'OpenSesame' sends the marker
from the following item and is therefore sequence-ordered. We recommend a
photodiode check in either case, but the 'PsychoPy' path removes the dominant
source of onset jitter. Notably, neither the matching nor the script generation
imports 'PsychoPy' or any serial library: generation only writes text, and the
presentation software is needed solely at run time.

## Reproducibility and FAIR compliance

'lexsync' ships a `renv` lockfile for R and pinned requirements for Python, sets
seeds from a single configuration file, and carries a 'CITATION.cff', an MIT
licence for the code and CC BY-SA terms for the bundled corpus derivatives with
full attribution. The R package passes `R CMD check --as-cran`; the Python package
builds a wheel and a source distribution that pass `twine check`. Both ship unit
test suites, including a mock-engine test that confirms the generated 'PsychoPy'
script writes its onset trigger on the word's first flip and a structural
validator for the generated `.osexp`.

# Worked examples

We illustrate 'lexsync' with four demonstrations: a high- versus low-frequency
contrast and a dense- versus sparse-neighbourhood contrast, each in English and
in Spanish, drawn from ten-thousand-word lexica. The frequency contrast matches
items on length, neighbourhood density and OLD20; the neighbourhood contrast
matches on length and frequency. Each demonstration was produced by both the R
and the Python engine. All tables and figures in this section are computed at
render time from the generated output.

```{r summary-table}
#| label: tbl-summary
#| tbl-cap: "Each demonstration, its manipulated dimension and the absolute standardised mean difference of the manipulated dimension and of the most-different controlled dimension. The manipulation is large in every case; the controlled dimensions remain small."
summ <- do.call(rbind, lapply(names(controlled_map), function(des) {
  x <- comps[comps$design == des, ]
  man <- x[x$role == "Manipulated", ]
  ctrl <- x[x$role == "Controlled", ]
  data.frame(
    Demonstration = des,
    `Manipulated dimension` = as.character(dim_labels[man$dimension]),
    `|d| manipulated` = round(man$absd, 2),
    `largest |d| controlled` = round(max(ctrl$absd), 2),
    check.names = FALSE, row.names = NULL
  )
}))
knitr::kable(summ, align = c("l", "l", "r", "r"))
```

@tbl-summary shows that the manipulated dimension is strongly separated in every
demonstration. The control dimensions, by contrast, are closely equated.
@fig-balance plots the absolute standardised mean difference of each dimension
against the anchor; the manipulated dimension stands far to the right of the
0.5-standard-deviation equivalence bound (dashed line) while the controlled
dimensions cluster near zero.

```{r balance-figure}
#| label: fig-balance
#| fig-cap: "Absolute standardised mean difference of each dimension against the anchor condition, in each demonstration. The dashed line is the 0.5-SD equivalence bound. Manipulated dimensions (orange) lie far to the right; matched dimensions (green) cluster near zero. In the neighbourhood contrasts OLD20 is free to co-vary (purple), reflecting its known correlation with Coltheart's N."
#| fig-width: 9
#| fig-height: 5.5
if (have_ggplot) {
  library(ggplot2)
  ggplot(comps, aes(x = absd, y = dim_label, colour = role)) +
    geom_vline(xintercept = 0.5, linetype = "dashed", colour = "grey60") +
    geom_segment(aes(x = 0, xend = absd, yend = dim_label), linewidth = 0.7) +
    geom_point(size = 3.1) +
    facet_wrap(~design, ncol = 2) +
    scale_colour_manual(values = c("Manipulated" = "#d95f02", "Controlled" = "#1b9e77",
                                   "Free (co-varying)" = "#7570b3"), drop = FALSE) +
    labs(x = "|Cohen's d| versus the anchor condition", y = NULL, colour = NULL) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "top", panel.grid.minor = element_blank(),
          strip.text = element_text(face = "bold"))
} else {
  knitr::kable(comps[, c("design", "dim_label", "cohens_d", "role")])
}
```

The per-condition descriptive statistics make the control concrete. In the
English frequency contrast (@tbl-desc-freq) frequency is separated by roughly one
and a half Zipf units while length, neighbourhood density and OLD20 are almost
identical across conditions. In the English neighbourhood contrast
(@tbl-desc-nd) neighbourhood density is separated by an order of magnitude with
length and frequency held constant; OLD20, which is mathematically related to N,
moves with the manipulation rather than being held constant, as expected.

```{r desc-freq}
#| label: tbl-desc-freq
#| tbl-cap: "Descriptive statistics per condition, English frequency contrast (R engine)."
knitr::kable(pretty_desc("en_freqcontrast_english"))
```

```{r desc-nd}
#| label: tbl-desc-nd
#| tbl-cap: "Descriptive statistics per condition, English neighbourhood-density contrast (R engine)."
knitr::kable(pretty_desc("en_ndensity_english"))
```

Finally, @tbl-parity reports the cross-engine parity: for each demonstration we
compare the set of (word, condition) pairs selected by the R engine with those
selected by the Python engine. The overlap is complete in every case, confirming
that the two independent implementations realise the same deterministic
procedure.

```{r parity-table}
#| label: tbl-parity
#| tbl-cap: "Cross-engine parity: Jaccard overlap between the stimuli selected by the R and Python engines."
parity <- sapply(designs, function(d) {
  r <- read_out(file.path("stimuli", paste0(d, "_stimuli_R.csv")))
  p <- read_out(file.path("stimuli", paste0(d, "_stimuli_py.csv")))
  rs <- paste(r$word, r$condition)
  ps <- paste(p$word, p$condition)
  length(intersect(rs, ps)) / length(union(rs, ps))
})
knitr::kable(data.frame(Demonstration = labels[designs],
                        `Jaccard overlap` = round(parity, 3),
                        check.names = FALSE),
             row.names = FALSE)
```

# Discussion

'lexsync' unifies steps that are usually performed in separate tools and separate
languages. Its contribution is integration rather than a single new capability:
multidimensional matching exists in 'LexOPS' and 'LIBRA', and precise
presentation exists in 'PsychoPy' and 'OpenSesame', but the path from a
multilingual corpus to a hardware-timed, cross-platform experiment has not before
been available as a single installable, dual-language tool. Two features are
worth emphasising. First, the deterministic matcher gives bit-for-bit agreement
between the R and Python engines, which means a laboratory can adopt 'lexsync' in
whichever ecosystem it prefers without fear of divergent stimuli. Second, the
'PsychoPy' export upgrades the sequence-ordered triggers of the original workflow
to genuine flip-locking, a concrete methodological improvement for ERP research.

Several limitations should be stated plainly. The orthographic-neighbourhood
measures are defined for alphabetic scripts and require adaptation for
logographic languages. The 'OpenSesame' export is sequence-ordered rather than
flip-locked, reflecting the platform's design; a photodiode check remains
advisable for any onset-critical study. The demonstrations use ten-thousand-word
lexica, which are adequate for illustration but smaller than a researcher would
use for a definitive neighbourhood calculation; larger corpora are fetched on
demand through the registry. Future work includes additional presentation targets,
a graphical interface and richer lexical dimensions such as concreteness and
age of acquisition, each of which slots into the existing module structure.

# Availability and open practices

'lexsync' is available as an R package and a Python package, with a shared corpus
registry and the reproducible source of this manuscript. The code is released
under the MIT licence and the bundled corpus derivatives under CC BY-SA 4.0, with
full attribution and retrieval dates recorded alongside the data. The repository
URL in the package metadata is a placeholder pending public release.

## Extending lexsync

In keeping with the design, extensions have well-defined insertion points. To add
a lexical dimension, edit `compute`-style functions in `querying` (`add_neighbourhood`
and the dimension table in `config/schema.yaml`) in both packages. To add a
presentation target, add an `export_<target>()` function in `scripting`. To add a
corpus or a language, add an entry to `corpora/registry.yaml`; no code change is
required for SUBTLEX-family or 'wordfreq' sources.

# References

::: {#refs}
:::
