---
title: "lexsync: A cross-platform pipeline for multidimensional lexical optimisation and hardware-timed experiment generation"
author:
  - name: Pablo Bernabeu
    affiliations:
      - name: Department of Education, University of Oxford
        city: Oxford
        country: United Kingdom
    email: pablo.bernabeu@education.ox.ac.uk
date: today
abstract: |
  Psycholinguistic stimulus preparation is fragmented across tools that match
  words, tools that present them and the separate R and Python ecosystems in
  which those tools live. We present 'lexsync', a research-grade toolkit,
  released as a 'CRAN'-ready R package and a structurally identical
  'PyPI'-ready Python package, that unifies the whole path from corpus to
  laboratory. From a word-frequency corpus in any of dozens of languages
  'lexsync' selects stimuli that are matched in parallel across several lexical
  dimensions (length, frequency, orthographic neighbourhood density and OLD20),
  counterbalances them, reports the realised match quality and then generates
  ready-to-run 'PsychoPy' and 'OpenSesame' experiment scripts whose hardware
  triggers for electroencephalography are time-locked to stimulus onset in both
  targets. The matching is deterministic, so the two engines select identical
  matched stimuli; we
  demonstrate
  this on five worked examples in English, Spanish and Mandarin Chinese, the last
  of which shows that the workflow extends to a logographic writing system. The
  generation step
  requires no presentation software and no hardware, which keeps the whole
  workflow reproducible. 'lexsync' generalises the FAIR artificial-grammar
  workflow of an existing longitudinal study and is distributed under open
  licences with a machine-readable corpus registry.
bibliography: references.bib
csl: apa.csl
format:
  html:
    toc: false
    number-sections: true
    df-print: kable
    embed-resources: true
  pdf:
    toc: false
    number-sections: true
    df-print: kable
    pdf-engine: xelatex
    papersize: a4
    geometry: margin=1in
  docx: default
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
have_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
find_out <- function(p) {
  for (base in c("../output", "output")) {
    f <- file.path(base, p)
    if (file.exists(f)) return(f)
  }
  stop(sprintf("lexsync manuscript: output file not found: %s", p))
}
read_out <- function(p) read.csv(find_out(p), stringsAsFactors = FALSE, fileEncoding = "UTF-8")

designs <- c("en_freqcontrast_english", "en_ndensity_english",
             "es_freqcontrast_spanish", "es_ndensity_spanish",
             "zh_freqcontrast_chinese")
labels <- c(
  en_freqcontrast_english = "English: frequency contrast",
  en_ndensity_english     = "English: neighbourhood contrast",
  es_freqcontrast_spanish = "Spanish: frequency contrast",
  es_ndensity_spanish     = "Spanish: neighbourhood contrast",
  zh_freqcontrast_chinese = "Chinese: frequency contrast"
)
comps <- do.call(rbind, lapply(designs, function(d) {
  x <- read_out(file.path("reports", paste0(d, "_comparisons_R.csv")))
  x$design <- labels[[d]]
  x
}))

# Readable dimension labels and the role of each dimension in each design. The
# length label is script-neutral because the demonstrations span alphabetic
# (letters) and logographic (characters) writing systems.
dim_labels <- c(frequency = "Frequency (Zipf)", length = "Length",
                n_density = "Neighbourhood N", old20 = "OLD20")
find_cfg <- function(f) {
  for (b in c("../config", "config")) {
    p <- file.path(b, f); if (file.exists(p)) return(p)
  }
  stop(f)
}
cfg_files <- c(
  "English: frequency contrast"     = "design_en_freqcontrast.yaml",
  "English: neighbourhood contrast" = "design_en_ndensity.yaml",
  "Spanish: frequency contrast"     = "design_es_freqcontrast.yaml",
  "Spanish: neighbourhood contrast" = "design_es_ndensity.yaml",
  "Chinese: frequency contrast"     = "design_zh_freqcontrast.yaml"
)
controlled_map <- lapply(cfg_files, function(f) unlist(yaml::read_yaml(find_cfg(f))$match_on))
names(controlled_map) <- names(cfg_files)

comps$absd <- abs(comps$cohens_d)
comps <- do.call(rbind, lapply(split(comps, comps$design), function(x) {
  ctrl <- controlled_map[[x$design[1]]]
  x$role <- ifelse(x$dimension %in% ctrl, "Controlled", "Free (co-varying)")
  free <- which(x$role != "Controlled")
  if (length(free)) x$role[free[which.max(x$absd[free])]] <- "Manipulated"
  x
}))
comps$dim_label <- factor(dim_labels[comps$dimension],
                          levels = c("Frequency (Zipf)", "Length", "Neighbourhood N", "OLD20"))
comps$role <- factor(comps$role, levels = c("Manipulated", "Controlled", "Free (co-varying)"))

pretty_desc <- function(d) {
  x <- read_out(file.path("reports", paste0(d, "_descriptives_R.csv")))
  x$group <- gsub("_", " ", x$group)
  x$dimension <- as.character(dim_labels[x$dimension])
  names(x) <- c("Condition", "Dimension", "n", "Mean", "SD", "Min", "Median", "Max")
  x
}
```

# Introduction

The credibility of psycholinguistic and neurolinguistic research rests on
carefully controlled stimuli and on precise timing. Both are laborious to
achieve, and the tools that support them are scattered. A researcher who wishes
to contrast, say, high- and low-frequency words while equating length and
orthographic neighbourhood typically matches the items in one program, lays out
the trial structure in another and writes the trigger code that aligns each word
with the electroencephalography (EEG) recording by hand. Each hand-off is an
opportunity for error and a barrier to reuse. The 'FAIR' guiding principles
[@wilkinson2016fair] — that research objects be findable, accessible,
interoperable and reusable — apply as much to this preparatory machinery as to
the data it ultimately yields, yet stimulus pipelines are seldom shared in a
form that others can run end to end.

Two families of tool address parts of the problem. The first matches or equates
word stimuli on lexical properties. 'LexOPS' [@taylor2020lexops] generates
controlled word lists for arbitrary factorial designs, with both per-dimension
tolerances and weighted multivariate Euclidean matching; 'LIBRA'
[@lintz2021libra] equates lists across conditions using a resampling algorithm;
and 'LASTU' [@itkonen2024lastu] provides searchable access to a language-specific
lexicon. These tools are mature and widely used, but they end at the stimulus
list: none generates the experiment that presents it, and most are tied to a
single language and to one programming ecosystem. The second family builds and
runs experiments. 'PsychoPy' [@peirce2019psychopy2] and 'OpenSesame'
[@mathot2012opensesame] offer precise spatial and temporal control over stimulus
presentation and support the hardware triggers that ERP research requires, but
they take the stimulus set as given and do nothing to construct or to balance it.

The gap, then, is not a missing capability but a missing integration. No single,
installable framework spans many-language corpus access, parallel
multidimensional matching, counterbalancing and the automated generation of
hardware-timed presentation scripts, and none does so across both the R and the
Python ecosystems and both major open presentation platforms. A second, subtler
gap concerns timing. In the artificial-grammar workflow that motivated this work
[@gonzalezalonso2025l3], EEG markers were sent from an experiment item that
followed the word, so that only the first marker was even approximately aligned
with stimulus onset; the others were sequence-ordered. This is a common pattern
and an avoidable source of jitter.

We present 'lexsync', a toolkit that fills these gaps; both presentation targets time-lock the marker to stimulus onset (see below). It is released as two
independent but structurally identical packages — a 'CRAN'-ready R package and a
'PyPI'-ready Python package — that share module names, function names and
behaviour. From a corpus 'lexsync' selects matched stimuli, counterbalances and
documents them, then writes a 'PsychoPy' script in which each onset trigger is
bound to the exact screen flip on which the word appears, using
`win.callOnFlip`, together with a complete 'OpenSesame' experiment. Crucially the
matching is deterministic, so the R and Python engines select identical matched stimuli.
'lexsync' generalises the materials workflow of @gonzalezalonso2025l3 into a
reusable, FAIR-compliant tool.

# The lexsync packages

## Architecture and cross-language design

Each package is organised into the same modules: `querying` (load a lexicon and
compute lexical dimensions), `matching` (the constraint engine), `counterbalancing`,
`validation` (match-quality reporting), `scripting` (experiment generation),
`logging` (the automated run log) and `corpora` (the registry). A single
orchestrator, `run_pipeline`, drives them in an explicit loop over designs and
languages, so that adding a language or a design is a matter of supplying a
configuration file and a lexicon rather than writing code. The two packages are
deliberately lean: the R package depends only on 'readr', 'yaml', 'stringdist'
and 'jsonlite' beyond base R, and the Python package on 'pandas', 'numpy',
'PyYAML', 'rapidfuzz' and 'scipy'.

## A many-language corpus registry

Stimuli are only as good as the corpus behind them. 'lexsync' reads a
machine-readable registry that records, for each supported corpus, its language,
access URL, licence and citation. Two connectors are provided. The first exposes
the curated SUBTLEX family and the 'openlexicon' collection — for example
SUBTLEX-UK for British English [@vanheuven2014subtlexuk], SUBTLEX-ESP for Spanish
[@cuetos2011subtlexesp] and SUBTLEX-US [@brysbaert2009subtlexus] — each
individually citable and redistributable under CC BY-SA terms. The second is the
'wordfreq' package [@speer2022wordfreq], which yields frequencies for roughly
forty languages through a single dependency. The demonstrations below build their
English, Spanish and Mandarin Chinese lexica with 'wordfreq', which aggregates
subtitle and web corpora including the SUBTLEX lists; the registry documents the
curated corpora — among them SUBTLEX-CH for Chinese [@cai2010subtlexch] — as
alternatives. Small example lexica are bundled inside each package for offline use
and testing, while full corpora are fetched on demand into a user cache, keeping
the installed packages within distribution size limits.

## The multidimensional matching algorithm

For every word 'lexsync' records its length, its frequency and two
orthographic-neighbourhood measures: Coltheart's N, the number of words of the
same length differing by a single letter [@coltheart1977access], and OLD20, the
mean Levenshtein distance to the twenty nearest neighbours [@yarkoni2008old20].
The neighbourhood measures are computed transparently from the lexicon rather
than taken on trust.

Matching proceeds in two deterministic stages. First, a tolerance pre-filter
keeps only candidates that fall within *M* ± *k* · *SD* of the anchor condition on
each control dimension, generalising the windowed selection of the original
workflow. Second, every remaining condition is matched to the anchor item by
item: each dimension is standardised, the standardised Euclidean distance from
each candidate to the target item is computed and the nearest unused candidate is
assigned, with ties broken by a stable, locale-independent key. Because no random
number generator is used and string comparisons are performed in byte order in
both engines, the R and Python packages select an identical matched selection
from identical input — a parity we verify below.

The per-anchor procedure fixes one condition and matches the others to it, which
is ideal when the manipulated dimension is statistically independent of the
controls. When the manipulation is intrinsically correlated with a control
dimension — as orthographic neighbourhood density is with word length and
frequency — fixing one condition first can leave a residual difference on the
control, because the anchor items need not have close counterparts in the other
condition. For such designs 'lexsync' offers a `joint` matching method, selected
with a single line in the design configuration. Rather than privileging one
condition, it scores every cross-condition pair on the standardised control
dimensions and greedily retains the *n* cheapest disjoint pairs, so only items
that have a close counterpart in the opposite condition are kept. The two
conditions are thereby equated on the controls even when the manipulation is
confounded with them, at the cost of drawing from a slightly larger candidate
pool. The method is as deterministic and cross-engine identical as the per-anchor
one (costs are rounded before the byte-order tie-break), and it is the method used
for the neighbourhood-density demonstrations below.

## Counterbalancing and reporting

Matched items are assigned to presentation lists and a reproducible trial order,
and a participant table crosses any counterbalancing factors. 'lexsync' then
writes a match-quality report and an automated, comprehensive run log recording
each stage, its parameters, the seed, package and corpus provenance and a
fingerprint of every file written.

The report is deliberately built around effect sizes rather than significance
tests. For each controlled dimension it gives the per-condition descriptive
statistics and the standardised mean difference (Cohen's *d*) between conditions,
together with a 90% confidence interval on that difference. This ordering is a
considered response to a common error in nuisance control: treating a
non-significant difference test as proof that a variable is matched
[@sassenhagen2016nuisance]. A non-significant test is not evidence of equivalence,
and whether a real difference reaches significance depends on the number of items,
so the conclusion "matched" can be manufactured by using few items. The
standardised mean difference is, by contrast, a direct measure of the realised
imbalance, and its confidence interval makes the sampling uncertainty — and hence
the dependence on the number of items — explicit: with few items the interval is
wide, so a small point estimate cannot be over-read, and the upper limit of the
interval on \|*d*\| is the largest imbalance still consistent with the chosen
stimuli. The report also includes a two one-sided tests (TOST) equivalence test
[@lakens2017equivalence] as a complementary, bound-referenced verdict; the 90%
interval is exactly the interval whose lying inside the bound corresponds to a
TOST at the .05 level, so the two are coherent and the interval is the more
transparent of the pair. Together they make the realised control explicit, in the
spirit of the original workflow's balance checks, and the run log makes a run
auditable.

## Hardware time-locking

The 'scripting' module emits two presentation targets from the same matched
stimuli. The 'PsychoPy' script binds each onset trigger to the stimulus flip:

```python
win.callOnFlip(port.setData, target_trigger)   # written on the onset flip
for frame in range(WORD_DURATION_FRAMES):
    word_stim.draw()
    win.flip()
    if frame == RESET_AFTER_FRAMES:
        win.callOnFlip(port.setData, 0)
```

A mock port prints the codes when no parallel-port driver is present, so the
script runs end to end with no hardware. The 'OpenSesame' export is a complete,
plain-text `.osexp`, generated programmatically so that its tab-sensitive format
is always valid and modelled on a proven-working experiment from
@gonzalezalonso2025l3. In it the word is drawn and shown from an inline script,
and the marker is sent immediately after `show()` returns, supporting both a
parallel-port and a serial backend with a test-mode fallback. Because `show()`
blocks until the display refresh, the marker is written right after the verified
flip, so both targets are time-locked to stimulus onset: 'PsychoPy' binds the
marker to the flip with `win.callOnFlip`, and 'OpenSesame' writes it immediately
after the refresh-blocking `show()`, the platform's recommended method. We
recommend a photodiode check in either case. Notably, neither the matching nor the script generation
imports 'PsychoPy' or any serial library: generation only writes text, and the
presentation software is needed solely at run time.

## Reproducibility and FAIR compliance

'lexsync' declares its dependencies in a `DESCRIPTION` file for R and pinned requirements for Python, sets
seeds from a single configuration file, and carries a 'CITATION.cff', an MIT
licence for the code and CC BY-SA terms for the bundled corpus derivatives with
full attribution. The R package passes `R CMD check --as-cran`; the Python package
builds a wheel and a source distribution that pass `twine check`. Both ship unit
test suites, including a mock-engine test that confirms the generated 'PsychoPy'
script writes its onset trigger on the word's first flip and a structural
validator for the generated `.osexp`.

# Worked examples

We illustrate 'lexsync' with five demonstrations. Four are alphabetic: a high-
versus low-frequency contrast and a dense- versus sparse-neighbourhood contrast,
each in English and in Spanish, drawn from thirty-thousand-word lexica. The
frequency contrast matches items on length, neighbourhood density and OLD20; the
neighbourhood contrast matches on length and frequency. The fifth is a
high- versus low-frequency contrast in Mandarin Chinese, included to show that the
pipeline generalises to a logographic writing system (see *Generalisation to a
logographic script*, below). Each demonstration was produced by both the R and the
Python engine. All tables and figures in this section are computed at render time
from the generated output.

```{r summary-table}
#| label: tbl-summary
#| tbl-cap: "*Manipulated and Controlled Standardised Mean Differences by Demonstration.* *Note.* Each demonstration, its manipulated dimension and the absolute standardised mean difference (Cohen's *d*) of the manipulated dimension and of the most-different controlled dimension. The manipulation is large in every case; the controlled dimensions remain small."
summ <- do.call(rbind, lapply(names(controlled_map), function(des) {
  x <- comps[comps$design == des, ]
  man <- x[x$role == "Manipulated", ]
  ctrl <- x[x$role == "Controlled", ]
  data.frame(
    Demonstration = des,
    `Manipulated dimension` = as.character(dim_labels[man$dimension]),
    `|d| manipulated` = round(man$absd, 2),
    `largest |d| controlled` = round(max(ctrl$absd), 2),
    check.names = FALSE, row.names = NULL
  )
}))
knitr::kable(summ, align = c("l", "l", "r", "r"))
```

@tbl-summary shows that the manipulated dimension is strongly separated in every
demonstration. The control dimensions, by contrast, show standardised mean
differences well below the 0.5-standard-deviation bound.
@fig-balance plots the absolute standardised mean difference of each dimension
against the anchor; the manipulated dimension stands far to the right of the
0.5-standard-deviation reference bound (dashed line) while the controlled
dimensions cluster near zero.

```{r balance-figure}
#| label: fig-balance
#| fig-cap: "*Absolute Standardised Mean Differences by Dimension and Demonstration.* *Note.* Absolute standardised mean difference (Cohen's *d*) of each dimension against the anchor condition. The dashed line is the 0.5-*SD* reference bound (the smallest effect of interest for the equivalence test). Manipulated dimensions (orange) lie far to the right; matched dimensions (green) cluster near zero. In the neighbourhood contrasts OLD20 is free to co-vary (purple), reflecting its known correlation with Coltheart's N."
#| fig-width: 9
#| fig-height: 5.5
if (have_ggplot) {
  library(ggplot2)
  ggplot(comps, aes(x = absd, y = dim_label, colour = role)) +
    geom_vline(xintercept = 0.5, linetype = "dashed", colour = "grey60") +
    geom_segment(aes(x = 0, xend = absd, yend = dim_label), linewidth = 0.7) +
    geom_point(size = 3.1) +
    facet_wrap(~design, ncol = 2) +
    scale_colour_manual(values = c("Manipulated" = "#d95f02", "Controlled" = "#1b9e77",
                                   "Free (co-varying)" = "#7570b3"), drop = FALSE) +
    labs(x = expression("|Cohen's " * italic(d) * "| versus the anchor condition"), y = NULL, colour = NULL) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "top", panel.grid.minor = element_blank(),
          strip.text = element_text(face = "bold"))
} else {
  knitr::kable(comps[, c("design", "dim_label", "cohens_d", "role")])
}
```

The realised control is best read from the effect sizes and their intervals.
@tbl-tost lists, for the controlled dimensions of each demonstration, Cohen's *d*
against the anchor, its 90% confidence interval and the complementary TOST verdict
at the 0.5-*SD* bound.

```{r tost-table}
#| label: tbl-tost
#| tbl-cap: "*Realised Control on the Controlled Dimensions: Effect Sizes With Confidence Intervals and the Complementary Equivalence Test.* *Note.* For each demonstration's controlled (matched) dimensions: the standardised mean difference (Cohen's *d*) against the anchor, its 90% confidence interval, and the two one-sided tests (TOST) *p* value and verdict at the 0.5-*SD* bound. The 90% interval is the one coherent with a .05-level TOST. R engine."
tost <- comps[comps$role == "Controlled",
              c("design", "dim_label", "cohens_d", "d_ci_low", "d_ci_high", "tost_p", "equivalent")]
tost$ci <- sprintf("[%.2f, %.2f]", tost$d_ci_low, tost$d_ci_high)
tab <- tost[, c("design", "dim_label", "cohens_d", "ci", "tost_p", "equivalent")]
names(tab) <- c("Demonstration", "Dimension", "Cohen's d", "90% CI", "TOST p", "Equivalent")
knitr::kable(tab, digits = 3, row.names = FALSE)
```

In every demonstration each controlled dimension's 90% confidence interval lies
within the 0.5-*SD* bound (*n* = 80 per condition), so the realised imbalance is
small even at the interval's far end — not merely non-significant. The TOST agrees,
declaring each dimension equivalent at the .05 level, but the interval is the more
informative statement because it would widen, and the claim would weaken, were the
item count reduced. The neighbourhood contrasts are the demanding case:
neighbourhood density is intrinsically correlated with word length and frequency,
so a per-anchor match leaves a residual on length (Cohen's *d* ≈ 0.25). The
`joint` matching method removes it: by selecting the best-matched dense–sparse
pairs jointly it equates length and frequency almost exactly (*d* ≈ 0.00 on both,
with intervals tightly around zero) while the density contrast remains an order of
magnitude larger (*d* > 2). The confound is thus a solved engineering problem
rather than an inherent limit on the design, and 'lexsync' reports the realised
control transparently — effect size and interval first, equivalence test as a
bound-referenced complement [@lakens2017equivalence; @sassenhagen2016nuisance].

The per-condition descriptive statistics make the control concrete. In the
English frequency contrast (@tbl-desc-freq) frequency is separated by roughly one
and a half Zipf units while length, neighbourhood density and OLD20 are almost
identical across conditions. In the English neighbourhood contrast
(@tbl-desc-nd) neighbourhood density is separated by an order of magnitude with
length and frequency held constant; OLD20, which is mathematically related to N,
moves with the manipulation rather than being held constant, as expected.

```{r desc-freq}
#| label: tbl-desc-freq
#| tbl-cap: "*Descriptive Statistics per Condition: English Frequency Contrast.* *Note.* R engine."
knitr::kable(pretty_desc("en_freqcontrast_english"))
```

```{r desc-nd}
#| label: tbl-desc-nd
#| tbl-cap: "*Descriptive Statistics per Condition: English Neighbourhood-Density Contrast.* *Note.* R engine."
knitr::kable(pretty_desc("en_ndensity_english"))
```

## Generalisation to a logographic script

The orthographic-neighbourhood measures operate on character strings, so the
pipeline is not tied to alphabetic writing. To show this concretely, the fifth
demonstration is a high- versus low-frequency contrast in Mandarin Chinese. Its
lexicon is derived from the same engine as the others — `wordfreq`'s Chinese data
[@speer2022wordfreq] — and the two conditions are two-character words, the
dominant word form in Chinese. Here 'length' is the number of characters, and
Coltheart's N and OLD20 are computed over characters rather than letters; we refer
to them as character-level neighbourhood measures. Adding the demonstration
required no change to the matching, counterbalancing or scripting code: a design
configuration names the Chinese lexicon and a glyph-complete font, and the
existing machinery does the rest.

@tbl-desc-zh shows the result. Frequency is separated by about one and a half Zipf
units (Cohen's *d* ≈ 6), while character-length is held constant at two and the
character-level neighbourhood measures are equated across conditions (each with a
90% confidence interval on *d* lying within the 0.5-*SD* bound, and
correspondingly TOST-equivalent). Chinese two-character words have dense
orthographic neighbourhoods — tens of single-character substitutions each, an
order of magnitude more than the alphabetic words above — and 'lexsync' equates
this dense, skewed distribution as readily as the sparse alphabetic one. The
generated 'PsychoPy' and 'OpenSesame' experiments set a CJK font, so the
characters render directly; the selection is byte-identical across the R and
Python engines, exactly as for the alphabetic demonstrations.

```{r desc-zh}
#| label: tbl-desc-zh
#| tbl-cap: "*Descriptive Statistics per Condition: Mandarin Chinese Frequency Contrast.* *Note.* Length is in characters; neighbourhood N and OLD20 are character-level measures. R engine."
knitr::kable(pretty_desc("zh_freqcontrast_chinese"))
```

```{r parity-setup}
#| include: false
parity_counts <- vapply(designs, function(d) {
  r <- read_out(file.path("stimuli", paste0(d, "_stimuli_R.csv")))
  p <- read_out(file.path("stimuli", paste0(d, "_stimuli_py.csv")))
  rs <- paste(r$word, r$condition); ps <- paste(p$word, p$condition)
  c(identical = length(intersect(rs, ps)), total = length(union(rs, ps)))
}, numeric(2))
n_identical <- sum(parity_counts["identical", ])
n_total <- sum(parity_counts["total", ])
n_match <- isTRUE(n_identical == n_total)
```

Finally, because the matcher is deterministic the two engines select the same
matched (word, condition) pairs rather than merely similar ones (the seeded trial
order differs, as the R and Python random-number generators do). Across the five
demonstrations
`r if (n_match) "all" else n_identical` `r n_total` of the selected
(word, condition) pairs were identical between the R and the Python engine, and
every statistic they report agrees to the precision shown. @tbl-parity makes this
concrete for the English frequency contrast: the standardised mean difference of
each dimension, computed independently in the two languages, coincides to the
precision shown.

```{r parity-table}
#| label: tbl-parity
#| tbl-cap: "*Cross-Engine Parity for the English Frequency Contrast.* *Note.* Each dimension's standardised mean difference (Cohen's *d*) against the anchor, computed independently by the R and Python engines. The columns are identical to three decimal places."
rcmp <- read_out("reports/en_freqcontrast_english_comparisons_R.csv")
pcmp <- read_out("reports/en_freqcontrast_english_comparisons_py.csv")
ptab <- data.frame(
  Dimension = as.character(dim_labels[rcmp$dimension]),
  `Cohen's d (R engine)` = round(rcmp$cohens_d, 3),
  `Cohen's d (Python engine)` = round(pcmp$cohens_d[match(rcmp$dimension, pcmp$dimension)], 3),
  check.names = FALSE, row.names = NULL
)
knitr::kable(ptab, align = c("l", "r", "r"))
```

# Discussion

'lexsync' unifies steps that are usually performed in separate tools and separate
languages. Its contribution is integration rather than a single new capability:
multidimensional matching exists in 'LexOPS' and 'LIBRA', and precise
presentation exists in 'PsychoPy' and 'OpenSesame', but the path from a
multilingual corpus to a hardware-timed, cross-platform experiment has not before
been available as a single installable, dual-language tool. Two features are
worth emphasising. First, the deterministic matcher selects identical matched
(word, condition) sets in the R and Python engines, with the reported match
statistics agreeing to the precision shown; only the seeded trial order, which
depends on each language's random-number generator, differs. A laboratory can
therefore adopt 'lexsync' in whichever ecosystem it prefers without fear of
divergent stimulus selection. Second, both
presentation targets upgrade the sequence-ordered triggers of the original
workflow to onset-aligned markers — flip-locked in 'PsychoPy', and written
immediately after the refresh-blocking `show()` in 'OpenSesame' — a concrete
methodological improvement for ERP research.

Some limitations remain. The Chinese demonstration shows that the pipeline
generalises to a logographic script — corpus access, dimension derivation,
matching, counterbalancing and CJK-font script generation all work unchanged —
but the neighbourhood measures it equates there (Coltheart's N and OLD20) are
computed at the character level. They do not capture the sub-character structure
that also shapes Chinese reading, such as radical, stroke-count and
phonetic-radical information. Such measures are not intrinsic to the method; they
enter as additional script-specific dimensions through the same extensible
dimension system, and supplying them is left to future work. A photodiode check
remains advisable for any onset-critical study, as for any software trigger.
Future work includes
additional presentation targets, a graphical interface and richer lexical
dimensions such as concreteness and age of acquisition, each of which slots into
the existing module structure.

# Availability and open practices

'lexsync' is available as an R package and a Python package, with a shared corpus
registry and the reproducible source of this manuscript. The code is released
under the MIT licence and the bundled corpus derivatives under CC BY-SA 4.0, with
full attribution and retrieval dates recorded alongside the data. The repository
URL in the package metadata is a placeholder pending public release.

## Extending lexsync

In keeping with the design, extensions have well-defined insertion points. To add
a lexical dimension, edit `compute`-style functions in `querying` (`add_neighbourhood`
and the dimension table in `config/schema.yaml`) in both packages. To add a
presentation target, add an `export_<target>()` function in `scripting`. To add a
corpus or a language, add an entry to `corpora/registry.yaml`; no code change is
required for SUBTLEX-family or 'wordfreq' sources.

{{< pagebreak >}}

# References

::: {#refs}
:::
