---
title: "lexsync: A cross-platform pipeline for multidimensional lexical optimisation and hardware-timed experiment generation"
author:
  - name: Pablo Bernabeu
    affiliations:
      - name: Department of Education, University of Oxford
        city: Oxford
        country: United Kingdom
    email: pablo.bernabeu@education.ox.ac.uk
date: today
abstract: |
  Psycholinguistic stimulus preparation is fragmented across tools that match
  words, tools that present them and the separate R and Python ecosystems in
  which those tools live. We present 'lexsync', a toolkit that unifies the
  whole path from corpus to laboratory, released as a 'CRAN'-ready R package
  and a structurally identical 'PyPI'-ready Python package. From a
  word-frequency corpus in any of dozens of languages
  'lexsync' selects stimuli that are matched in parallel across several lexical
  dimensions (length, frequency, orthographic neighbourhood density and OLD20),
  counterbalances them, reports the realised match quality and then generates
  ready-to-run 'PsychoPy', 'OpenSesame' and browser-based 'jsPsych' experiments.
  The two laboratory targets time-lock their electroencephalography triggers to
  stimulus onset. The trial is described declaratively as a sequence of events, so the
  same engine builds factorial word studies, lexical decision with generated
  pseudowords, priming and self-paced reading from configuration rather than code.
  The default matching methods, the pseudoword generation and the trial order
  are deterministic, so the two engines produce byte-identical experiments. We
  demonstrate this on 21 worked examples in English, Spanish and Mandarin
  Chinese (across word, pseudoword, paired and sentence stimuli), one of which
  shows that the workflow extends to a logographic writing system, while three
  reproduce published designs from the literature. The generation step
  requires no presentation software and no hardware, which keeps the whole
  workflow reproducible. 'lexsync' is general-purpose and treats the stimulus
  set as a reproducible artefact. It is distributed under open licences and
  includes a machine-readable corpus registry.
bibliography: references.bib
csl: apa.csl
format:
  html:
    toc: false
    number-sections: true
    df-print: kable
    embed-resources: true
  pdf:
    toc: false
    number-sections: true
    df-print: kable
    pdf-engine: xelatex
    papersize: a4
    geometry: margin=1in
  docx: default
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
have_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
find_out <- function(p) {
  for (base in c("../output", "output")) {
    f <- file.path(base, p)
    if (file.exists(f)) return(f)
  }
  stop(sprintf("lexsync manuscript: output file not found: %s", p))
}
read_out <- function(p) read.csv(find_out(p), stringsAsFactors = FALSE, fileEncoding = "UTF-8")

designs <- c("en_freqcontrast_english", "en_ndensity_english",
             "es_freqcontrast_spanish", "es_ndensity_spanish",
             "zh_freqcontrast_chinese")
labels <- c(
  en_freqcontrast_english = "English: frequency contrast",
  en_ndensity_english     = "English: neighbourhood contrast",
  es_freqcontrast_spanish = "Spanish: frequency contrast",
  es_ndensity_spanish     = "Spanish: neighbourhood contrast",
  zh_freqcontrast_chinese = "Chinese: frequency contrast"
)
comps <- do.call(rbind, lapply(designs, function(d) {
  x <- read_out(file.path("reports", paste0(d, "_comparisons_R.csv")))
  x$design <- labels[[d]]
  x
}))

# Readable dimension labels and the role of each dimension in each design. The
# length label is script-neutral because the demonstrations span alphabetic
# (letters) and logographic (characters) writing systems.
dim_labels <- c(frequency = "Frequency (Zipf)", length = "Length",
                n_density = "Neighbourhood N", old20 = "OLD20")
find_cfg <- function(f) {
  for (b in c("../config", "config")) {
    p <- file.path(b, f); if (file.exists(p)) return(p)
  }
  stop(f)
}
cfg_files <- c(
  "English: frequency contrast"     = "design_en_freqcontrast.yaml",
  "English: neighbourhood contrast" = "design_en_ndensity.yaml",
  "Spanish: frequency contrast"     = "design_es_freqcontrast.yaml",
  "Spanish: neighbourhood contrast" = "design_es_ndensity.yaml",
  "Chinese: frequency contrast"     = "design_zh_freqcontrast.yaml"
)
controlled_map <- lapply(cfg_files, function(f) unlist(yaml::read_yaml(find_cfg(f))$match_on))
names(controlled_map) <- names(cfg_files)

comps$absd <- abs(comps$cohens_d)
comps <- do.call(rbind, lapply(split(comps, comps$design), function(x) {
  ctrl <- controlled_map[[x$design[1]]]
  x$role <- ifelse(x$dimension %in% ctrl, "Controlled", "Free (co-varying)")
  free <- which(x$role != "Controlled")
  if (length(free)) x$role[free[which.max(x$absd[free])]] <- "Manipulated"
  x
}))
comps$dim_label <- factor(dim_labels[comps$dimension],
                          levels = c("Frequency (Zipf)", "Length", "Neighbourhood N", "OLD20"))
comps$role <- factor(comps$role, levels = c("Manipulated", "Controlled", "Free (co-varying)"))

pretty_desc <- function(d) {
  x <- read_out(file.path("reports", paste0(d, "_descriptives_R.csv")))
  x$group <- gsub("_", " ", x$group)
  x$dimension <- as.character(dim_labels[x$dimension])
  names(x) <- c("Condition", "Dimension", "n", "Mean", "SD", "Min", "Median", "Max")
  x
}
```

# Introduction

The credibility of psycholinguistic and neurolinguistic research rests on
carefully controlled stimuli and on precise timing. Both are laborious to
achieve, and the tools that support them are scattered. A researcher who wishes
to contrast, say, high- and low-frequency words while equating length and
orthographic neighbourhood typically matches the items in one program, lays out
the trial structure in another and writes the trigger code that aligns each word
with the electroencephalography (EEG) recording by hand. Each hand-off is an
opportunity for error and a barrier to reuse. The 'FAIR' guiding principles
[@wilkinson2016fair] (that research objects be findable, accessible,
interoperable and reusable) apply as much to this preparatory machinery as to
the data it ultimately yields, yet stimulus pipelines are seldom shared in a
form that others can run end to end.

This preparatory machinery is where a quiet reproducibility problem lives.
A survey of 600 randomly sampled linguistics articles found materials, data
and protocols each shared in fewer than 10% of them, with no noticeable rise
between 2008/2009 and 2018/2019 [@bochynska2023reproducible]. Yet
materials are a precondition both for replicating a result and for judging how far
it generalises. The gap compounds an analytic one. Quantitative phonetic and
psycholinguistic work affords many undocumented analytic choices, from how a
measure is quantified to how a model is specified, and exploiting that
flexibility inflates false positives [@roettger2019degrees] and yields widely
divergent conclusions from the same data [@coretta2023multidimensional].
Transparent, runnable, shareable materials are part of the remedy this
literature calls for [@roettger2019emergent].

There is also a statistical reason to take materials seriously. The items in an
experiment are a sample from a larger population of possible items, and a
conclusion is meant to hold for that population, not merely for the words actually
shown. Treating items as a fixed rather than a random factor inflates Type I error
(the language-as-fixed-effect fallacy [@clark1973language]), and recent work
recasts the same point as a *generalizability crisis*: researchers routinely model
participants as a random factor but not stimuli, and so claim more than their
designs license [@yarkoni2020generalizability]. A tool that makes the construction
of a matched, balanced item set explicit, reproducible and shareable is therefore
part of the methodological infrastructure that careful inference requires.

Two families of tool address parts of the problem. The first matches or equates
word stimuli on lexical properties. 'LexOPS' [@taylor2020lexops] generates
controlled word lists for arbitrary factorial designs, with both per-dimension
tolerances and weighted multivariate Euclidean matching; 'LIBRA'
[@lintz2021libra] equates lists across conditions using a genetic (evolutionary)
algorithm;
and 'LASTU' [@itkonen2024lastu] provides searchable access to a language-specific
lexicon. These tools are mature and widely used, but they end at the stimulus
list. None generates the experiment that presents it, and none is available in
both the R and the Python ecosystems. Their language coverage follows the
lexical database each ships with rather than the method itself, and both
'LexOPS' and 'LIBRA' will work from any suitably formatted dataset, in any
language for which the user can supply the variables. The second family builds and
runs experiments. 'PsychoPy' [@peirce2019psychopy2] and 'OpenSesame'
[@mathot2012opensesame] offer precise spatial and temporal control over stimulus
presentation and support the hardware triggers that ERP research requires, but
they take the stimulus set as given and do nothing to construct or to balance it.

The gap, then, is one of integration rather than of capability. No single,
installable framework spans many-language corpus access, parallel
multidimensional matching, counterbalancing and the automated generation of
hardware-timed presentation scripts, and none does so across both the R and the
Python ecosystems and both major open presentation platforms. A second, subtler
gap concerns timing. In one longitudinal EEG study of artificial language
learning, for instance [@gonzalezalonso2025l3], EEG markers were sent from an
experiment item that followed the word, so that only the first marker was even
approximately aligned with stimulus onset and the others were sequence-ordered.
Their timing then followed the durations of the preceding events rather than the
flip that put the word on screen, a dependency that binding the trigger to the
flip itself removes.

We present 'lexsync', a toolkit that fills these gaps. It is released as two
independent but structurally identical packages (a 'CRAN'-ready R package and a
'PyPI'-ready Python package) that share module names, function names and
behaviour. From a corpus 'lexsync' selects matched stimuli, counterbalances and
documents them. It then writes a 'PsychoPy' script in which each onset trigger is
bound, using `win.callOnFlip`, to the exact screen flip on which the critical
stimulus appears, together with a complete 'OpenSesame' experiment. The
default matching methods and the trial ordering are deterministic, so the R and
Python engines select identical materials, order them identically and write
byte-identical experiment files. The trial itself is described declaratively. A
design sets out an ordered sequence of trial *events*, and each presentation
backend renders that sequence, so the same machinery produces a factorial word
study, a lexical-decision task with generated pseudowords, a primed design or a
self-paced reading study from configuration rather than new code. 'lexsync' is
thus a reusable, FAIR-compliant tool for many psycholinguistic paradigms and
languages rather than a single study's script. The longitudinal study above is
one of 21 worked examples, and one of the three that rebuild a published
design.

# The lexsync packages

## Architecture and cross-language design

Each package is organised into the same modules: `querying` (load a lexicon or an
item table and compute lexical dimensions), `matching` (the constraint engine),
`generation` (deterministic pseudoword generation), `paradigms` (the trial-event
registry), `counterbalancing`, `validation` (match-quality reporting), `scripting`
(experiment generation), `datasheet` (the materials datasheet and pre-registration
template), `logging` (the automated run log) and `corpora` (the registry). A
single orchestrator, `run_pipeline`, drives them for one design, and `run_all`
applies it in an explicit loop over every design configuration in a directory,
so that adding a language, a design or a paradigm is a matter of supplying a
configuration file (and, where relevant, a lexicon or item table) rather than
writing code. The two packages are deliberately lean: the R package depends only
on 'readr', 'yaml', 'stringdist', 'stringi', 'jsonlite' and 'digest' beyond base
R, and the Python package on 'pandas', 'numpy', 'PyYAML', 'rapidfuzz' and
'scipy'.

## A many-language corpus registry

The quality of the matched stimuli depends on the corpus from which they are
drawn. 'lexsync' reads a machine-readable registry that records, for each corpus
it knows about, its language, licence, citation, verification status and the
column names that map the source file onto the package schema. Two connectors
are provided. The first reads the delimited files of the curated SUBTLEX family
and the 'openlexicon' collection (for example SUBTLEX-ESP for Spanish
[@cuetos2011subtlexesp] and SUBTLEX-US for American English
[@brysbaert2009subtlexus]), each individually citable and made freely available
by its authors subject to attribution, on terms equivalent to CC BY-SA. No entry
in that family currently registers a directly addressable delimited file, so the
registry marks the corpora the connector cannot read and gives a landing page
instead. SUBTLEX-UK [@vanheuven2014subtlexuk], for one, is published only as
zipped archives, so the researcher downloads the file once and passes it to
`load_lexicon()`. The second connector is the 'wordfreq' package
[@speer2022wordfreq], which supplies frequencies through a single dependency for
the thirty languages the registry currently admits, among them Greek, Russian,
Arabic, Hebrew, Hindi, Chinese, Japanese and Korean. The package itself covers
around forty, and the rest are enabled by adding their language code to the
registry entry. The demonstrations below build their English, Spanish and
Mandarin Chinese lexica with 'wordfreq', which aggregates subtitle and web
corpora and, for some languages including English and Chinese, the corresponding
SUBTLEX lists. The registry documents the curated corpora (among them SUBTLEX-CH
for Chinese [@cai2010subtlexch]) as alternatives. Small example lexica are
bundled inside each package for offline use and testing, which keeps the
installed packages within distribution size limits. A lexicon for any of the
admitted 'wordfreq' languages is derived on demand into a user cache, an
ingestion step that runs in Python, and the R package reads the result as an
ordinary corpus.

## The multidimensional matching algorithm

For every word 'lexsync' records its length, its frequency and two
orthographic-neighbourhood measures: Coltheart's N, the number of words of the
same length differing by a single letter [@coltheart1977access], and OLD20, the
mean Levenshtein distance to the twenty nearest neighbours [@yarkoni2008old20].
Both are computed transparently from the lexicon rather than taken on trust, and
the reference set is the full loaded lexicon rather than the filtered candidate
pool the items are drawn from. A word is excluded from its own neighbourhood,
and where fewer than twenty other forms are available OLD20 averages over as
many as there are. The dimension set is open. It also includes an orthographic
syllable estimate (the number of maximal vowel runs, counting *y* and the
accented Latin vowels as vowels) and a bigram-frequency measure, the mean over a
word's adjacent letter bigrams of that bigram's share of all bigram occurrences
in the lexicon, which approximates phonotactic probability from word forms
rather than from frequency-weighted tokens. Externally-sourced norms
(concreteness, age of acquisition, valence) can be joined on through a
norm-table connector so the matcher equates on them too. A sixth corpus
demonstration puts two of the added dimensions to use, a frequency contrast
matched on length, syllable count and bigram frequency. Both engines select the
same items for it.

A design declares its candidate pool by filtering the lexicon on any recorded
dimension, and then defines each condition by a range or a set of permitted
values on one or more dimensions, so each condition reaches the matcher as its
own subpool. The first condition listed is the anchor. Its items are not drawn
at random, and they are not taken from one end of the subpool either. The
subpool is sorted by the first dimension the condition is defined on, with ties
broken by the UTF-8 bytes of the word, and the *n* items are taken at evenly
spaced ranks along it, so the anchor spans the subpool rather than clustering at
one end.

Matching then proceeds in two deterministic stages. First, a tolerance
pre-filter keeps only candidates that fall within *M* ± *k* · *SD* of the anchor
condition on each control dimension, generalising the windowed selection used by
@gonzalezalonso2025l3. The schema sets *k* to 1 for frequency and to 2 for
length, Coltheart's N and OLD20, and any dimension it does not name falls back
to 2. A design file may override these values, as the reproduction of that study
below does. If a condition holds fewer candidates inside its window than the
design asks for, the window is dropped for that condition and the match is made
over its whole subpool, so a window set too narrow costs matching precision
rather than the run. The matcher stops with an error only when the subpool
itself is too small, or too incomplete on the matched dimensions, to fill the
design, and the remedy is then to widen the pool filters or to ask for fewer
items per condition. The relaxation is reported on the console as it happens,
though it is not yet recorded in the datasheet, so a design with deliberately
narrow windows should be run with that output in view. The continuous selector
treats its control windows the same way.

Second, every remaining condition is matched to the anchor item by item. Each
dimension is standardised on the mean and sample standard deviation of the whole
candidate pool rather than of a single condition, so one scaling serves every
comparison. The standardised Euclidean distance from each candidate to the
target item is then computed and rounded to nine decimal places, which absorbs
the last-bit differences between the two language runtimes, and the nearest
candidate not yet taken is assigned, with ties broken by the UTF-8 bytes of the
word. A word assigned in one condition is withdrawn from every later one, so no
word appears twice. Because no random number generator is used and string
comparisons are performed in byte order in both engines, the R and Python
packages select an identical matched selection from identical input, a parity we
verify below.

Two further matching methods can be selected when a design needs them. A
covariance-aware (Mahalanobis) distance replaces the standardised Euclidean metric
with one weighted by the inverse of the control dimensions' correlation matrix, so
it down-weights dimensions that are correlated, such as length and orthographic
neighbourhood, rather than counting their shared variance more than once
[@rubin1980bias; @stuart2010matching]. A globally optimal assignment, for a
two-condition design, minimises the total distance over the whole pairing instead
of taking the locally cheapest pair at each step [@gu1993comparison;
@hansen2006optimal]. Both rely on a linear-algebra routine (a matrix inverse and
an assignment solver respectively) whose last bits differ between the R and
Python backends, so on these two methods the engines select equivalent but not
byte-identical materials; each datasheet records which case applies. The optimal
method also carries a dependency the other methods do not. In R its assignment
solver comes from the suggested package 'clue', installed on demand rather than
alongside 'lexsync', and in Python from 'SciPy', which the package already
requires.

A design may also be declared *continuous* rather than dichotomised. Instead of
splitting a variable into conditions and matching, 'lexsync' selects a set that
spans the variable's range evenly while holding the control dimensions
near-constant, so the variable stays continuous and is analysed by regression or a
mixed model rather than a between-condition contrast. This is the recommended
alternative to matched dichotomies, which lose power and admit selection
artefacts [@kuperman2015virtual; @libennowell2019danger]. For a continuous
design the datasheet reports the realised span, the predictor-control
correlations and a suggested regression model.

The per-anchor procedure fixes one condition and matches the others to it, which
is ideal when the manipulated dimension is statistically independent of the
controls. When the manipulation is intrinsically correlated with a control
dimension (as orthographic neighbourhood density is with word length and
frequency), fixing one condition first can leave a residual difference on the
control, because the anchor items need not have close counterparts in the other
condition. For such designs 'lexsync' offers a `joint` matching method, selected
with a single line in the design configuration. Rather than privileging one
condition, it scores cross-condition pairs on the standardised control
dimensions and greedily retains the *n* cheapest disjoint pairs, so only items
that have a close counterpart in the opposite condition are kept. To keep the
pair matrix tractable, each condition's candidate set is first reduced, in turn,
to the 1,200 items closest to the other condition's centroid in standardised
space. In both neighbourhood demonstrations below at least one condition's
candidate set is larger than that, so the reduction is active in them. The two
conditions are thereby equated on the controls even when the manipulation is
confounded with them, at the cost of drawing from a slightly larger candidate
pool. The method is as deterministic and cross-engine identical as the per-anchor
one (costs are rounded before the byte-order tie-break), and it is the method used
for the neighbourhood-density demonstrations below.

## Counterbalancing and reporting

Matched items are assigned to presentation lists and a reproducible trial order,
and a participant table crosses any counterbalancing factors. The trial order
within each list comes from a seeded, keyed-hash shuffle: each trial is ranked
by the SHA-256 digest of the string `seed|replicate|list|set|condition`, a
tuple that identifies it uniquely under either counterbalancing recipe.
Distinct inputs to SHA-256 behave as independent uniform draws, so ranking by
the digest realises a seeded random permutation, yet the permutation is a pure
function of the design. A different seed gives a different order, and the same
seed gives the same order, byte for byte, from both engines on any platform.
Nothing in the ordering consults the machine, since the digest and the byte-order
sort that ranks trials by it are fixed by the specification rather than by the
local toolchain. That is checked rather than assumed. The parity suite runs on
Linux, macOS and Windows, and on every push and pull request continuous
integration generates the experiments with each engine in turn and fails if any
file differs. No random-number generator is involved anywhere in the package, so
there is no generator state to save or restore. The permutation is uniform,
which makes it a legitimate randomisation and not merely a reproducible one.
Over a range of seeds each item's mean serial position sits at the centre of its
list, and each item reaches every position about equally often. The test suite
checks the centring and that every item reaches first place at some seed. A
design may also request *resampling*: several disjoint matched item sets are
drawn, no item reused across them, so a study can run a different item sample per
participant group or demonstrate an effect across samples rather than resting on
a single fixed set of items. Generalising beyond the particular items shown is a
long-standing concern [@clark1973language; @yarkoni2020generalizability].
'lexsync' then
writes a match-quality report and an automated, comprehensive run log recording
each stage, its parameters, the seed, package and corpus provenance and a
fingerprint of every file written.

The report is deliberately built around effect sizes rather than significance
tests. For each controlled dimension it gives the per-condition descriptive
statistics and the standardised mean difference (Cohen's *d*, the anchor
condition minus the comparison condition over their pooled standard deviation),
together with a 90% confidence interval on that difference. The interval is the
ordinary *t* interval on the raw mean difference, re-expressed in pooled standard
deviations. Its half-width is the .95 quantile of *t* on *n*~1~ + *n*~2~ − 2
degrees of freedom, times the square root of 1/*n*~1~ + 1/*n*~2~, so it carries
the sampling uncertainty of the two means but not that of the pooled standard
deviation, and it is therefore not the non-central *t* interval sometimes used
for *d*. This ordering is a
considered response to a common error in nuisance control: treating a
non-significant difference test as proof that a variable is matched
[@sassenhagen2016nuisance]. A non-significant test is not evidence of equivalence,
and whether a real difference reaches significance depends on the number of items,
so the conclusion "matched" can be manufactured by using few items. The
standardised mean difference is, by contrast, a direct measure of the realised
imbalance, and its confidence interval makes the sampling uncertainty (and hence
the dependence on the number of items) explicit. With few items the interval is
wide, so a small point estimate cannot be over-read, and the upper limit of the
interval on \|*d*\| is the largest imbalance still consistent with the chosen
stimuli. The report also includes a two one-sided tests (TOST) equivalence test
[@lakens2017equivalence] as a complementary verdict against the schema's bound of
0.5 pooled standard deviations. A 90% interval falling entirely within that bound
is exactly the condition under which a TOST at the .05 level declares
equivalence, so the two are coherent, and the interval is the more transparent of
the pair. Together they make the realised control explicit, in the spirit of the
balance checks that study-specific pipelines perform by hand, and the run log
makes a run auditable.

Finally, each design is accompanied by a *materials datasheet*, a machine-readable
JSON record and a human-readable Markdown summary of its provenance. The record
covers the corpus or item table and its checksum, the dimensions and how they were
computed, the selection or matching method and its parameters, the realised control
(the effect sizes and their intervals, a variance-ratio check on the spread, and the
equivalence verdicts), the counterbalancing
recipe, and the seed, software versions and file checksums needed to reproduce the
materials exactly. It records, in addition, the size of the candidate pool each
condition was selected from, so that item selection is auditable rather than
opaque, and a suggested crossed mixed-model formula that treats items as a random
factor [@clark1973language], with which it fills the pre-registration analysis
plan. The datasheet also emits an auto-filled Methods paragraph and a
pre-registration template with its Materials section completed. This is a direct
response to the finding that experimental materials are rarely shared in a usable
form [@bochynska2023reproducible]: the materials, the procedure and the record
of how they were made travel together.

## Hardware time-locking

Of the three presentation targets the 'scripting' module emits from the same
matched stimuli, two drive laboratory hardware. The browser-based third is
described below. In the 'PsychoPy' script a single routine, `show_frames`, draws
every timed event and binds any onset trigger to the flip on which the stimulus
first appears:

```python
def show_frames(win, stim, frames, port, trigger):
    if trigger is not None:
        win.callOnFlip(port.setData, trigger)
    for f in range(frames):
        if stim is not None:
            stim.draw()
        win.flip()
        if trigger is not None and f == RESET_AFTER_FRAMES:
            win.callOnFlip(port.setData, 0)
    if trigger is not None and frames <= RESET_AFTER_FRAMES:
        port.setData(0)
```

The code is held for `triggers.trigger_hold_ms`, 50 ms by default, which the
generated script converts to whole flips against the refresh rate it measures, and
the reset is itself bound to a flip. Holding it in milliseconds rather than in
frames is what keeps the pulse above the recorder's minimum on a fast display. An
event that does not last beyond that window clears the port directly as it ends.

A mock port prints the codes when no parallel-port driver is present, so the
script runs end to end with no hardware. The 'OpenSesame' export is a complete,
plain-text `.osexp`. It is generated programmatically, so its tab-sensitive
format is always valid, and it is modelled on a working experiment reported by
@gonzalezalonso2025l3. In it the word is drawn and shown from an inline script,
and the marker is sent as soon as `show()` returns, supporting both a
parallel-port and a serial backend with a test-mode fallback. Because `show()`
blocks until the display refresh, that marker follows the verified flip, which is
the method the platform recommends. Both laboratory targets are therefore
time-locked to stimulus onset, though we recommend a photodiode check in either
case. Neither the matching nor the script generation imports 'PsychoPy' or any
serial library. Generation only writes text, and the presentation software is
needed solely at run time.

## Paradigm generality: the trial-event model

A study-specific script typically supports one kind of trial (a fixation cross, a
single word and a response), hard-coded into the experiment. 'lexsync' replaces
that fixed structure with a declarative one. A trial is an ordered list of *events*,
each a small record naming its type (a fixation, a piece of text, a mask, a blank
interval, a region-by-region passage, a response or a comprehension question), its
content (a literal such as a fixation cross, or a reference to a field of the
trial such as the target word or the sentence), its duration, and an optional EEG
trigger that may be locked to the event's onset. A *paradigm* is a named
default sequence of such events together with the trial fields it requires; a
design either selects a paradigm or supplies its own event list. All three
presentation backends iterate the same list. The 'PsychoPy' script interprets it
at run time, the 'OpenSesame' generator emits one block per event, and the
'jsPsych' generator builds a browser timeline from it. Adding a paradigm is
therefore a matter of configuration rather than backend code, and all three
targets present the same events in the same order. Every event length is declared
in milliseconds, which is the unit all three targets schedule, so a design
describes the same interval on any display. The 'OpenSesame' and 'jsPsych'
generators emit that duration directly. The 'PsychoPy' script measures the
display's refresh rate when it starts and converts each duration into the nearest
whole number of flips, so the stimulus still appears on a refresh and the onset
trigger is still bound to it, while the realised duration tracks the declared one
at any refresh rate. What remains is a quantisation of at most half a frame, which
the materials datasheet records. On the 'OpenSesame' target the onset code is also
held for 10 ms before it is reset, and that wait lengthens the event it marks by
the same amount. This concerns the presentation targets only. The generated files
themselves remain byte-identical between the R and Python engines, which emit the
same conversion.

The 'jsPsych' target is a single shareable HTML file that runs in any modern
browser, so anyone can reproduce the exact procedure online from the same
materials. The jsPsych library itself loads from a content delivery network.
Because a browser cannot drive a parallel port, each onset marker is recorded in
the trial's data rather than written to hardware, and online EEG synchronisation
would route it through WebSerial or a photodiode. Stimulus text is always written
as data into the conditions file the experiment reads, never interpolated into
the generated code, and any value containing a control character or exceeding a
length bound is rejected before writing, so a malformed item cannot corrupt a
script.

Items reach a design from one of three sources. The corpus source is the matching
engine described above. The *table* source loads a user-supplied set of items
(prime–target pairs, or sentences segmented into regions with a comprehension
question), validated against the paradigm's required fields and counterbalanced by
a Latin square that shows each item once per list in a rotated condition, so a
target is never repeated within a list. The *generate* source builds pseudowords
by one of two methods set in `items.generation.method`. The default,
`letter_substitution`, derives a length-matched, orthographically legal non-word
for each real word by changing the fewest letters such that every resulting
letter bigram is attested in the corpus and the form is neither a real word nor a
pseudoword already issued for the set, choosing among the legal candidates the
most bigram-plausible with a byte-order tie-break. Substituted letters are drawn
from a to z and the search stops after two changed positions, so the generator
reports a failure rather than drifting further from its base word. The
alternative, `subsyllabic`, splits each word into onset, nucleus and coda
constituents and swaps up to two thirds of them, rounded up, for attested
alternatives of the same role and length, taking codas first and onsets last, so
that the pseudoword keeps the syllabic structure of its base. Where no legal swap
is available for a word, that item falls back to letter substitution. Both are
deterministic orthographic approximations of 'Wuggy' [@keuleers2010wuggy] that
match length exactly and return byte-identical output from the R and the Python
engine.

## Reproducibility and FAIR compliance

'lexsync' declares its dependencies in a `DESCRIPTION` file for R and in pinned
requirements for Python, and it sets seeds from a single configuration file. The
distribution carries a 'CITATION.cff', an MIT licence for the code and CC BY-SA
terms for the bundled corpus derivatives, with full attribution. The R package
passes `R CMD check --as-cran`; the Python package builds a wheel and a source
distribution that pass `twine check`. Both ship unit test suites, including a
mock-engine test that confirms the generated 'PsychoPy' script writes its onset
trigger on the word's first flip and a structural validator for the generated
`.osexp`.

# Worked examples

We illustrate 'lexsync' with fifteen demonstrations. Six are corpus-matched word
contrasts, three exercise other paradigms, three put the analytic choices argued
for above into practice (a continuous design, a resampled one and a second
pseudoword generator) and the last three reproduce designs from the published
literature (see *Reproducing published designs*, below). The first four matched
contrasts cross two alphabetic languages: a high- versus low-frequency contrast
and a dense- versus sparse-neighbourhood contrast, each in English and in
Spanish, drawn from 30,000-word lexica. The frequency contrast matches items on
length, neighbourhood density and OLD20; the neighbourhood contrast matches on
length and frequency. The fifth is a high- versus low-frequency contrast in
Mandarin Chinese, included to show that the pipeline generalises to a logographic
writing system (see *Generalisation to a logographic script*, below). The sixth,
a further English contrast, matches on an orthographic syllable estimate and a
bigram-frequency measure in place of the two neighbourhood measures, exercising
the richer dimensions introduced with the matching algorithm above. The next
three (lexical decision, priming and self-paced reading) show the trial-event
model carrying word, pseudoword, paired and sentence stimuli (see *Beyond word
lists*, below). Each demonstration was produced by both the R and the Python
engine. All tables and figures in this section are computed at render time from
the generated output. @tbl-summary summarises the five frequency and
neighbourhood contrasts, for which a controlled-versus-manipulated report is
defined. The remaining demonstrations are described in their own sections.

```{r summary-table}
#| label: tbl-summary
#| tbl-cap: "*Manipulated and Controlled Standardised Mean Differences by Demonstration.* *Note.* Each demonstration, its manipulated dimension and the absolute standardised mean difference (Cohen's *d*) of the manipulated dimension and of the most-different controlled dimension. The manipulation is large in every case; the controlled dimensions remain small."
summ <- do.call(rbind, lapply(names(controlled_map), function(des) {
  x <- comps[comps$design == des, ]
  man <- x[x$role == "Manipulated", ]
  ctrl <- x[x$role == "Controlled", ]
  data.frame(
    Demonstration = des,
    `Manipulated dimension` = as.character(dim_labels[man$dimension]),
    `|d| manipulated` = round(man$absd, 2),
    `largest |d| controlled` = round(max(ctrl$absd), 2),
    check.names = FALSE, row.names = NULL
  )
}))
knitr::kable(summ, align = c("l", "l", "r", "r"))
```

@fig-balance plots the absolute standardised mean difference of each dimension
against the anchor condition. In every demonstration the manipulated dimension
stands far to the right of the 0.5-standard-deviation reference bound (dashed
line), while the controlled dimensions cluster near zero.

```{r balance-figure}
#| label: fig-balance
#| fig-cap: "*Absolute Standardised Mean Differences by Dimension and Demonstration.* *Note.* Absolute standardised mean difference (Cohen's *d*) of each dimension against the anchor condition. The dashed line is the 0.5-*SD* reference bound (the smallest effect of interest for the equivalence test). Manipulated dimensions (orange) lie far to the right; matched dimensions (green) cluster near zero. In the neighbourhood contrasts OLD20 is free to co-vary (purple), reflecting its known correlation with Coltheart's N."
#| fig-width: 9
#| fig-height: 5.5
if (have_ggplot) {
  library(ggplot2)
  ggplot(comps, aes(x = absd, y = dim_label, colour = role)) +
    geom_vline(xintercept = 0.5, linetype = "dashed", colour = "grey60") +
    geom_segment(aes(x = 0, xend = absd, yend = dim_label), linewidth = 0.7) +
    geom_point(size = 3.1) +
    facet_wrap(~design, ncol = 2) +
    scale_colour_manual(values = c("Manipulated" = "#d95f02", "Controlled" = "#1b9e77",
                                   "Free (co-varying)" = "#7570b3"), drop = FALSE) +
    labs(x = expression("|Cohen's " * italic(d) * "| versus the anchor condition"), y = NULL, colour = NULL) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "top", panel.grid.minor = element_blank(),
          strip.text = element_text(face = "bold"))
} else {
  knitr::kable(comps[, c("design", "dim_label", "cohens_d", "role")])
}
```

The realised control is best read from the effect sizes and their intervals.
@tbl-tost lists, for the controlled dimensions of each demonstration, Cohen's *d*
against the anchor, its 90% confidence interval and the complementary TOST verdict
at the 0.5-*SD* bound.

```{r tost-table}
#| label: tbl-tost
#| tbl-cap: "*Realised Control on the Controlled Dimensions: Effect Sizes With Confidence Intervals and the Complementary Equivalence Test.* *Note.* For each demonstration's controlled (matched) dimensions: the standardised mean difference (Cohen's *d*) against the anchor, its 90% confidence interval, and the two one-sided tests (TOST) *p* value and verdict at the 0.5-*SD* bound. The 90% interval is the one coherent with a .05-level TOST. R engine."
tost <- comps[comps$role == "Controlled",
              c("design", "dim_label", "cohens_d", "d_ci_low", "d_ci_high", "tost_p", "equivalent")]
tost$ci <- sprintf("[%.2f, %.2f]", tost$d_ci_low, tost$d_ci_high)
tab <- tost[, c("design", "dim_label", "cohens_d", "ci", "tost_p", "equivalent")]
names(tab) <- c("Demonstration", "Dimension", "Cohen's d", "90% CI", "TOST p", "Equivalent")
knitr::kable(tab, digits = 3, row.names = FALSE)
```

In each of these five demonstrations every controlled dimension's 90% confidence
interval lies within the 0.5-*SD* bound (*n* = 80 per condition), so the realised
imbalance is small even at the interval's far end, not merely non-significant.
The TOST agrees, declaring each dimension equivalent at the .05 level, but the
interval is the more informative statement because it would widen, and the claim
would weaken, were the item count reduced. The neighbourhood contrasts are the
demanding case:
neighbourhood density is intrinsically correlated with word length and frequency,
so a per-anchor match leaves a residual on length (Cohen's *d* ≈ 0.25). The
`joint` matching method removes it: by selecting the best-matched dense–sparse
pairs jointly it equates length and frequency almost exactly (*d* ≈ 0.00 on both,
with intervals tightly around zero) while the density contrast remains an order of
magnitude larger (*d* > 2). The confound is therefore something the matching
method can remove rather than an inherent limit on the design, and 'lexsync'
reports the realised control transparently, with effect size and interval first
and the equivalence test as a bound-referenced complement
[@lakens2017equivalence; @sassenhagen2016nuisance].

The per-condition descriptive statistics make the control concrete. In the
English frequency contrast (@tbl-desc-freq) frequency is separated by roughly one
and a half Zipf units while length, neighbourhood density and OLD20 are almost
identical across conditions. In the English neighbourhood contrast
(@tbl-desc-nd) neighbourhood density is separated by an order of magnitude with
length and frequency held constant; OLD20, which is mathematically related to N,
moves with the manipulation rather than being held constant, as expected.

```{r desc-freq}
#| label: tbl-desc-freq
#| tbl-cap: "*Descriptive Statistics per Condition: English Frequency Contrast.* *Note.* R engine."
knitr::kable(pretty_desc("en_freqcontrast_english"))
```

```{r desc-nd}
#| label: tbl-desc-nd
#| tbl-cap: "*Descriptive Statistics per Condition: English Neighbourhood-Density Contrast.* *Note.* R engine."
knitr::kable(pretty_desc("en_ndensity_english"))
```

## Generalisation to a logographic script

The orthographic-neighbourhood measures operate on character strings, so the
pipeline is not tied to alphabetic writing. To show this concretely, the fifth
demonstration is a high- versus low-frequency contrast in Mandarin Chinese. Its
lexicon comes from the same corpus package as the others (`wordfreq`'s Chinese data
[@speer2022wordfreq]), and the two conditions are two-character words, the
dominant word form in Chinese. Here 'length' is the number of characters, and
Coltheart's N and OLD20 are computed over characters rather than letters; we refer
to them as character-level neighbourhood measures. Adding the demonstration
required no change to the matching, counterbalancing or scripting code: a design
configuration names the Chinese lexicon and a glyph-complete font, and the
existing machinery does the rest.

@tbl-desc-zh shows the result. Frequency is separated by about one and a half Zipf
units (Cohen's *d* ≈ 6), while character-length is held constant at two and the
character-level neighbourhood measures are equated across conditions (each with a
90% confidence interval on *d* lying within the 0.5-*SD* bound, and
correspondingly TOST-equivalent). Chinese two-character words have dense
orthographic neighbourhoods (tens of single-character substitutions each, an
order of magnitude more than the alphabetic words above), and 'lexsync' equates
this dense, skewed distribution as readily as the sparse alphabetic one. The
generated 'PsychoPy' and 'OpenSesame' experiments set a CJK font, so the
characters render directly. The selection is byte-identical across the R and
Python engines, exactly as for the alphabetic demonstrations.

```{r desc-zh}
#| label: tbl-desc-zh
#| tbl-cap: "*Descriptive Statistics per Condition: Mandarin Chinese Frequency Contrast.* *Note.* Length is in characters; neighbourhood N and OLD20 are character-level measures. R engine."
knitr::kable(pretty_desc("zh_freqcontrast_chinese"))
```

## Cross-engine parity

```{r parity-setup}
#| include: false
parity_counts <- vapply(designs, function(d) {
  r <- read_out(file.path("stimuli", paste0(d, "_stimuli_R.csv")))
  p <- read_out(file.path("stimuli", paste0(d, "_stimuli_py.csv")))
  rs <- paste(r$word, r$condition); ps <- paste(p$word, p$condition)
  c(identical = length(intersect(rs, ps)), total = length(union(rs, ps)))
}, numeric(2))
n_identical <- sum(parity_counts["identical", ])
n_total <- sum(parity_counts["total", ])
n_match <- isTRUE(n_identical == n_total)
```

Because the matcher is deterministic the two engines select the same
matched (word, condition) pairs rather than merely similar ones, and because
the trial order is a keyed-hash function of the design they also place those
pairs in the same order. Across the five matched word contrasts
`r if (n_match) paste("all", n_total) else paste(n_identical, "of", n_total)` selected
(word, condition) pairs were identical between the R and the Python engine, and
every statistic they report agrees to the precision shown. That check runs when
this manuscript is rendered. The test suite makes a stricter one. For each of the
fifteen design configurations shipped with the package it regenerates the Python
selection and requires it to agree with the committed R selection on the selected
item, its condition and the set the matcher paired it into, and then on every
other column the two files share, value for value. Continuous integration goes
further and runs each engine into its own directory, comparing the two trees of
generated experiments recursively, so all 75 files (for every design, a
'PsychoPy' script and its conditions table, an 'OpenSesame' experiment and its
loop table, and a 'jsPsych' page) must agree byte for byte. The same job
regenerates the R selections first and fails if they have drifted from the
committed ones, so the comparison cannot pass against a stale reference. The
covariance-aware and optimal matching methods remain the exceptions, for the
reason given above. @tbl-parity makes this
concrete for the English frequency contrast: the standardised mean difference of
each dimension, computed independently in the two languages, coincides to the
precision shown.

```{r parity-table}
#| label: tbl-parity
#| tbl-cap: "*Cross-Engine Parity for the English Frequency Contrast.* *Note.* Each dimension's standardised mean difference (Cohen's *d*) against the anchor, computed independently by the R and Python engines. The columns are identical to three decimal places."
rcmp <- read_out("reports/en_freqcontrast_english_comparisons_R.csv")
pcmp <- read_out("reports/en_freqcontrast_english_comparisons_py.csv")
ptab <- data.frame(
  Dimension = as.character(dim_labels[rcmp$dimension]),
  `Cohen's d (R engine)` = round(rcmp$cohens_d, 3),
  `Cohen's d (Python engine)` = round(pcmp$cohens_d[match(rcmp$dimension, pcmp$dimension)], 3),
  check.names = FALSE, row.names = NULL
)
knitr::kable(ptab, align = c("l", "r", "r"))
```

## Beyond word lists: three paradigms

```{r paradigm-setup}
#| include: false
lex_cmp <- read_out("reports/en_lexdec_english_comparisons_R.csv")
lex_stim <- read_out("stimuli/en_lexdec_english_stimuli_R.csv")
prime_stim <- read_out("stimuli/en_priming_english_stimuli_R.csv")
spr_stim <- read_out("stimuli/en_spr_english_stimuli_R.csv")
n_lexdec <- sum(lex_stim$condition == "word")
lex_len_d <- lex_cmp$cohens_d[lex_cmp$dimension == "length"]
n_prime <- length(unique(prime_stim$set))
n_spr <- length(unique(spr_stim$set))
para_parity <- function(base, cols) {
  r <- read_out(paste0("stimuli/", base, "_stimuli_R.csv"))
  p <- read_out(paste0("stimuli/", base, "_stimuli_py.csv"))
  cols <- cols[cols %in% names(r) & cols %in% names(p)]
  setequal(do.call(paste, r[cols]), do.call(paste, p[cols]))
}
para_ok <- all(
  para_parity("en_lexdec_english", c("condition", "target")),
  para_parity("en_priming_english", c("list", "set", "condition", "prime", "target")),
  para_parity("en_spr_english", c("list", "set", "condition", "sentence")))
```

The same engine builds tasks that are not word lists. Three further demonstrations,
each in English, exercise a different stimulus structure and trial procedure
through the trial-event model. A lexical-decision design contrasts
`r n_lexdec` real words with `r n_lexdec` pseudowords generated to match them in
length; the generated non-words are orthographically legal and absent from the
corpus, and length is equated by construction (Cohen's *d* =
`r sprintf("%.2f", lex_len_d)`). A priming design draws `r n_prime`
prime–target pairs from an item table and contrasts related with unrelated primes,
its trial inserting a brief prime and a mask before the onset-locked target. A
self-paced reading design presents `r n_spr` sentences region by region, the
grammatical–ungrammatical manipulation falling on the onset-locked critical region
and followed by a comprehension question. Each design compiles from the same event
list to a runnable 'PsychoPy' script, a valid 'OpenSesame' experiment and a
browser-based 'jsPsych' experiment, and the R and Python engines
select `r if (para_ok) "identical" else "differing"` materials in every case,
including, for lexical decision, the very same deterministically generated
pseudowords.

## Selection modes for the intended analysis

```{r modes-setup}
#| include: false
cont_cmp <- read_out("reports/en_freqcontinuous_english_comparisons_R.csv")
cont_stim <- read_out("stimuli/en_freqcontinuous_english_stimuli_R.csv")
n_cont <- nrow(cont_stim)
cont_span <- sprintf("%.2f", unique(cont_cmp$predictor_span)[1])
cont_maxr <- sprintf("%.2f", max(abs(cont_cmp$pearson_r), na.rm = TRUE))
res_stim <- read_out("stimuli/en_resample_english_stimuli_R.csv")
res_cmp <- read_out("reports/en_resample_english_comparisons_R.csv")
n_resample <- nrow(res_stim)
n_res_sets <- length(unique(res_stim$replicate))
res_reuse <- n_resample != length(unique(res_stim$word))
res_maxd <- sprintf("%.2f",
                    max(abs(res_cmp$cohens_d[res_cmp$dimension != "frequency"])))
wug_stim <- read_out("stimuli/en_lexdec_wuggy_english_stimuli_R.csv")
wug_same_base <- setequal(lex_stim$target[lex_stim$condition == "word"],
                          wug_stim$target[wug_stim$condition == "word"])
wug_shared_pw <- length(intersect(
  lex_stim$target[lex_stim$condition == "pseudoword"],
  wug_stim$target[wug_stim$condition == "pseudoword"]))
modes_ok <- all(
  para_parity("en_freqcontinuous_english", c("word", "condition", "set")),
  para_parity("en_resample_english", c("replicate", "word", "condition", "set")),
  para_parity("en_lexdec_wuggy_english", c("target", "condition", "set")))
```

Three further demonstrations put the analytic choices argued for above into
practice, all drawn from the same English lexicon as the frequency contrast. The
continuous design abandons the high-low split. Its `r n_cont` items span
`r cont_span` Zipf units of frequency evenly while length, neighbourhood density
and OLD20 are held near-constant, and the largest correlation between the
predictor and any control dimension is \|*r*\| = `r cont_maxr`. The datasheet
accordingly proposes a mixed model with frequency entered continuously rather
than a between-condition contrast. The resampled design draws the same frequency
contrast into `r n_res_sets` disjoint matched sets, `r n_resample` items in
all, with `r if (res_reuse) "some words reused" else "no word reused"` across
sets, so a study can give a different item sample to each participant group and
treat items as the random factor they are [@clark1973language;
@yarkoni2020generalizability]. Control on length, neighbourhood density and
OLD20 holds across the pooled sets, the largest absolute standardised difference
on a controlled dimension being `r res_maxd`. The third repeats the
lexical-decision design with the `subsyllabic` generator in place of the
default, over `r if (wug_same_base) "the same" else "a different"` set of
`r n_lexdec` base words, so the two pseudoword methods can be compared like for
like. They have `r if (wug_shared_pw == 0) "no pseudoword" else "some pseudowords"`
in common. For all three the R and Python engines select
`r if (modes_ok) "identical" else "differing"` materials, and each datasheet
records the selection as byte-identical across the engines.

## Reproducing published designs

```{r repro-setup}
#| include: false
repro_comp <- function(d) {
  x <- read_out(file.path("reports", paste0(d, "_comparisons_R.csv")))
  x$absd <- abs(x$cohens_d)
  x
}
gen <- repro_comp("es_gender_repro_spanish")
ctrl_hi <- function(r) max(abs(c(r$d_ci_low, r$d_ci_high)))
gen_ctrl <- gen[gen$dimension %in% c("frequency", "length"), ]
gen_ctrl <- gen_ctrl[which.max(gen_ctrl$absd), ]
gen_ctrl_d  <- gen_ctrl$absd
gen_ctrl_hi <- ctrl_hi(gen_ctrl)
gen_n <- nrow(read_out(file.path("stimuli", "es_gender_repro_spanish_stimuli_R.csv")))
andr <- repro_comp("en_andrews_repro_english")
and_len <- andr[andr$dimension == "length", ]
and_len <- and_len[which.max(and_len$absd), ]
and_ctrl_d    <- and_len$absd
and_ctrl_hi   <- ctrl_hi(and_len)
and_ctrl_tost <- and_len$tost_p
and_freq_d <- max(andr$absd[andr$dimension == "frequency"])
and_nd_d   <- max(andr$absd[andr$dimension == "n_density"])
and_stim <- read_out(file.path("stimuli", "en_andrews_repro_english_stimuli_R.csv"))
and_n <- nrow(and_stim)
and_per_cell <- and_n / length(unique(and_stim$condition))
ras <- read_out(file.path("stimuli", "en_rastle_repro_english_stimuli_R.csv"))
ras_targets <- length(unique(ras$target))
ras_lists <- length(unique(ras$list))
```

The demonstrations so far are illustrative. To show that 'lexsync' reproduces
designs from the published literature, we recreated three studies that span the
workflow's range: a longitudinal EEG study of artificial language learning, a
classic factorial lexical-decision design and a masked-priming design. Each was
rebuilt to the original's matching *criteria* (its manipulated and controlled
dimensions and their tolerances) from a contemporary, redistributable corpus,
rather than by copying the authors' word lists. @tbl-repro summarises the three,
and the two engines reproduce each one identically.

```{r repro-table}
#| label: tbl-repro
#| tbl-cap: "*Three Published Designs Reproduced to Criteria.* *Note.* For each reproduction: the structure rebuilt, the number of generated items, and the largest standardised mean difference (Cohen's *d*) remaining on a controlled (matched) dimension, with in brackets the upper limit of its 90% confidence interval, that is the largest imbalance still consistent with the selected items. The manipulations are large by construction; for Andrews these are frequency and neighbourhood size, reported in the text. The Rastle design is a priming layout with no corpus-matched control dimension."
repro_tbl <- data.frame(
  Reproduction = c("@gonzalezalonso2025l3", "@andrews1989frequency",
                   "@rastle2004broth"),
  `Design rebuilt` = c("Gender assignment: feminine vs masculine, matched on frequency and length",
                       "Frequency x neighbourhood size: 2x2 factorial, cells matched on length",
                       "Masked priming: transparent / opaque / orthographic, vs unrelated baseline"),
  Items = c(gen_n, and_n, nrow(ras)),
  `Largest controlled |d| (upper 90% limit)` =
    c(sprintf("%.2f (%.2f)", gen_ctrl_d, gen_ctrl_hi),
      sprintf("%.2f (%.2f)", and_ctrl_d, and_ctrl_hi), "—"),
  check.names = FALSE, row.names = NULL
)
knitr::kable(repro_tbl, align = c("l", "l", "r", "r"))
```

@gonzalezalonso2025l3 administered a natural-language gender-assignment task in
which corpus nouns, half feminine and half masculine, were matched to the study's
nouns on word frequency (within mean ± *SD*/9) and number of letters (within
mean ± 2 *SD*). 'lexsync' reproduces this design
directly: the same per-dimension tolerance windows are declared in the design file
and applied by the matcher, and feminine and masculine words are equated on
frequency and length to a largest controlled Cohen's *d* of
`r sprintf("%.2f", gen_ctrl_d)` across `r gen_n` items. Because the bundled
Spanish corpus carries no lexical-gender field, gender is approximated by the
canonical ending (*-a*, *-o*) rather than looked up as the original did in EsPal
[@duchon2013espal]; the demonstration is therefore of the matched selection and
its realised control, the part 'lexsync' automates, not of the authors' exact
items.

@andrews1989frequency crossed word frequency with
orthographic neighbourhood size in a 2 × 2 design, run in lexical decision and in
naming, and found that a large neighbourhood facilitates low-frequency words
disproportionately. 'lexsync' rebuilds the factorial structure over short words,
equating all four cells on length (largest controlled \|*d*\| =
`r sprintf("%.2f", and_ctrl_d)`, upper 90% limit
`r sprintf("%.2f", and_ctrl_hi)`) while separating the manipulated dimensions
widely. With `r and_per_cell` items per cell the interval is wide, and length is
not shown equivalent at the 0.5-*SD* bound (TOST *p* =
`r sprintf("%.3f", and_ctrl_tost)`). The realised imbalance is small as a point
estimate, but at that cell size the upper limit is what the data support, and the
datasheet records it rather than letting the point estimate stand alone.
Frequency *d* reaches `r sprintf("%.1f", and_freq_d)` and neighbourhood *d*
reaches `r sprintf("%.1f", and_nd_d)`. Frequencies come from a contemporary corpus
rather than the Kučera–Francis counts of the original, so the realised bands differ
while the factorial design does not.

@rastle2004broth used masked priming to
dissociate three kinds of prime–target relationship, transparent morphological
(*cleaner*–CLEAN), opaque morphological (*corner*–CORN) and purely orthographic
(*brothel*–BROTH), each against an unrelated baseline, arguing for an early
morpho-orthographic segmentation stage blind to meaning. 'lexsync' expresses the
design with its priming paradigm: `r ras_targets` targets are each shown with a
related prime of their assigned type and an unrelated, length-matched control,
Latin-square counterbalanced across `r ras_lists` lists so a target appears once
per list, with the prime type carried as a between-items factor for analysis. This
is a representative item set following the prime-type structure of the original,
not its full materials.

# Discussion

'lexsync' unifies steps that are usually performed in separate tools and
separate languages. Its principal contribution is integration. Multidimensional
matching exists in 'LexOPS' and 'LIBRA', and precise presentation exists in
'PsychoPy' and 'OpenSesame', but the path from a multilingual corpus to a
hardware-timed, cross-platform experiment has not before been available as a
single installable, dual-language tool. Two features are worth emphasising.
First, the deterministic matcher selects identical matched (word, condition)
sets in the R and Python engines, with the reported match statistics agreeing to
the precision shown, and the keyed-hash trial order makes every generated
experiment file byte-identical between them; the sole exceptions remain the
covariance-aware (Mahalanobis) and optimal matching methods, whose
linear-algebra backends differ in their last bits. A laboratory can therefore
adopt 'lexsync' in whichever ecosystem it prefers without fear of divergent
stimulus selection. Second, both laboratory targets place every onset marker at
the stimulus flip (flip-locked in 'PsychoPy', and written immediately after the
refresh-blocking `show()` in 'OpenSesame'), rather than sending a trailing block
of markers in which only the first is even approximately aligned, a concrete
methodological improvement for ERP research.

Some limitations remain. The Chinese demonstration shows that the pipeline
generalises to a logographic script (corpus access, dimension derivation,
matching, counterbalancing and CJK-font script generation all work unchanged),
but the neighbourhood measures it equates there (Coltheart's N and OLD20) are
computed at the character level. They do not capture the sub-character structure
that also shapes Chinese reading, such as radical, stroke-count and
phonetic-radical information. Such measures are not intrinsic to the method; they
enter as additional script-specific dimensions through the same extensible
dimension system, and supplying them is left to future work. A photodiode check
remains advisable for any onset-critical study, as for any software trigger. The
example item tables for the paired and sentence paradigms are deliberately small,
intended to exercise and document the pipeline rather than to constitute validated
materials; researchers supply their own.

Several further extensions follow directly from the motivation and slot into the
existing module structure. The web target above could be joined by a 'PCIbex'
reading study and by client-side hardware synchronisation (WebSerial or a
photodiode patch), so that online studies inherit the same onset markers the
laboratory targets write. The dimension set already extends beyond orthography
to an orthographic syllable estimate and a corpus-derived bigram-frequency
(phonotactic-probability) measure, and it could grow further to predictability
dimensions such as n-gram or language-model surprisal. Semantic dimensions such
as concreteness, age of acquisition and valence can be joined on through the
norm-table connector. The norm tables themselves have to be obtained separately,
because their licences vary. Sampling matched
item sets per participant, rather than per study, would extend the resampling
already provided.

# Extending lexsync

In keeping with the design, extensions have well-defined insertion points. To add
a paradigm, add an entry to the registry in `paradigms` giving its default
event sequence, the trial fields it requires and its counterbalancing recipe, and
all three presentation backends then render it with no further code. To add a
lexical dimension, compute it in `querying` (alongside `add_neighbourhood`) and
list it in the dimension table in `config/schema.yaml`, in both packages. To add
a presentation target, add an `export_<target>()` function in `scripting` that
walks the same rendered event list. To add a corpus or language, add an entry to
`corpora/registry.yaml`, and no code change is required for SUBTLEX-family or
'wordfreq' sources.

# Availability and open practices

'lexsync' is available as an R package and a Python package, with a shared corpus
registry and the reproducible source of this manuscript. A browser application for
each engine (a Shiny app for the R package and a Streamlit app for the Python one)
lets a researcher assemble a design, run the verified pipeline and export the design
configuration together with the R, Python and command-line code that reproduces it,
so the interface itself yields shareable, reproducible artefacts. The code is released
under the MIT licence and the bundled corpus derivatives under CC BY-SA 4.0, with
full attribution and retrieval dates recorded alongside the data. The repository
is cited from the package metadata of both engines, from the citation file and
from the machine-readable software description, so a reader arrives at the same
source whichever record they follow.

{{< pagebreak >}}

# References

::: {#refs}
:::
