---
title: "Getting started with lexsync"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with lexsync}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`lexsync` selects stimuli that are matched in parallel across several lexical
dimensions and then generates the experiment scripts that present them. This
vignette runs a small high- versus low-frequency contrast on the bundled English
example lexicon.

The selection is deterministic, and it ships as two independent packages, one for
R and one for Python, which select byte-identical stimuli from the same lexicon
and design. No random number generator is involved: the engines agree because
every ordering is by byte and every tie is broken by a rule both apply. This holds
for the default matching methods; the optional `mahalanobis` and `optimal` methods
are equivalent but not byte-identical, because a covariance inverse and an
assignment solver differ in their last bits between the two platforms.

```{r setup}
library(lexsync)
schema <- yaml::read_yaml(system.file("extdata", "schema.yaml", package = "lexsync"))
lex <- load_lexicon(system.file("extdata", "en_example.csv", package = "lexsync"),
                    schema, language = "english")
nrow(lex)
```

## Define and run a design

A design names the conditions, the dimensions to match and the number of items,
together with the filters that narrow the lexicon to a candidate pool. The pool
is built first, exactly as the pipeline does, and matching then runs over it.

```{r design}
design <- list(
  name = "vignette_demo", language = "english", n_per_condition = 15,
  pool_filters = list(length = c(3, 7), frequency = c(3.8, 7)),
  conditions = list(
    list(name = "high", define_by = list(frequency = c(5.2, 7.0))),
    list(name = "low",  define_by = list(frequency = c(3.8, 4.4)))
  ),
  match_on = list("length", "n_density", "old20")
)
pool <- build_pool(lex, design$pool_filters)
stim <- match_stimuli(pool, design, schema)
head(stim[, c("word", "condition", "length", "frequency", "n_density", "old20")])
```

## Inspect the match quality

Matching always returns a set, so the report is what tells you whether the set is
any good. It belongs before anything is generated from the stimuli.

```{r report}
report <- match_report(stim, c("length", "frequency", "n_density", "old20"), schema)
knitr::kable(report$descriptives, caption = "Descriptive statistics per condition")
knitr::kable(report$comparisons,
             caption = "Standardised mean differences with 90% confidence intervals, plus the complementary TOST equivalence test")
```

The manipulated dimension (frequency) differs strongly, while the controlled
dimensions (length, neighbourhood density and OLD20) are closely equated. The
report leads with the standardised mean difference (`cohens_d`) and its 90%
confidence interval (`d_ci_low`, `d_ci_high`). A non-significant difference test is
not evidence of matching, and its outcome depends on the number of items. The
effect size and its interval, whose upper limit is the largest imbalance still
consistent with the stimuli, are therefore the primary summary. The TOST equivalence test
is reported alongside as a bound-referenced verdict.

## Generate the experiment scripts

The trial is described once, as a sequence of events, and each presentation target
renders that same description. One call therefore produces all three.

```{r export}
out <- file.path(tempdir(), "lexsync_demo")
dir.create(out, showWarnings = FALSE)
files <- export_experiments(stim, design, schema, out)
basename(unlist(files))
```

Three experiments are generated from one description of the trial. The PsychoPy
script binds each onset trigger to the stimulus flip with `win.callOnFlip`. The
OpenSesame `.osexp` draws and shows the word inside an `inline_script` and sends
the marker immediately after `show()` returns, which it does at the display
refresh. The jsPsych HTML runs the same procedure in a browser and records the
trigger codes in the trial data, since a browser cannot address a parallel port.
Neither the matching nor the script generation requires PsychoPy, OpenSesame or
any hardware to be installed.

## Where next

The pipeline does all of the above in one call. `run_pipeline()` on a design
configuration writes the stimuli, the reports, the three experiments, a materials
datasheet recording the provenance and realised control of the run, and a run log.

Three further articles go into depth. *Matching, dimensions and designs* covers
the lexical dimensions and their units, the four matching methods and when each
one suits, tolerance windows, continuous designs and how to read the validation
report. *Experiments, paradigms and EEG triggers* covers the trial-event model,
the paradigm registry, the three presentation targets and the flip-locked trigger
timing. *Reproducibility, parity and the materials datasheet* covers the
cross-engine guarantee, how it is achieved and where it stops.
