---
title: "Getting started with lexsync"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with lexsync}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`lexsync` selects stimuli that are matched in parallel across several lexical
dimensions and then generates the experiment scripts that present them. This
vignette runs a small high- versus low-frequency contrast on the bundled English
example lexicon.

```{r setup}
library(lexsync)
schema <- yaml::read_yaml(system.file("extdata", "schema.yaml", package = "lexsync"))
lex <- load_lexicon(system.file("extdata", "en_example.csv", package = "lexsync"),
                    schema, language = "english")
nrow(lex)
```

## Define and run a design

A design names the conditions, the dimensions to match and the number of items.

```{r design}
design <- list(
  name = "vignette_demo", language = "english", n_per_condition = 15,
  pool_filters = list(length = c(3, 7), frequency = c(3.8, 7)),
  conditions = list(
    list(name = "high", define_by = list(frequency = c(5.2, 7.0))),
    list(name = "low",  define_by = list(frequency = c(3.8, 4.4)))
  ),
  match_on = list("length", "n_density", "old20")
)
stim <- match_stimuli(lex, design, schema)
head(stim[, c("word", "condition", "length", "frequency", "n_density", "old20")])
```

## Inspect the match quality

```{r report}
report <- match_report(stim, c("length", "frequency", "n_density", "old20"), schema)
knitr::kable(report$descriptives, caption = "Descriptive statistics per condition")
knitr::kable(report$comparisons,
             caption = "Standardised mean differences with 90% confidence intervals, plus the complementary TOST equivalence test")
```

The manipulated dimension (frequency) differs strongly, while the controlled
dimensions (length, neighbourhood density and OLD20) are closely equated. The
report leads with the standardised mean difference (`cohens_d`) and its 90%
confidence interval (`d_ci_low`, `d_ci_high`): a non-significant difference test is
not evidence of matching, and its outcome depends on the number of items, so the
effect size and its interval — whose upper limit is the largest imbalance still
consistent with the stimuli — are the primary summary. The TOST equivalence test
is reported alongside as a bound-referenced verdict.

## Generate the experiment scripts

```{r export}
out <- file.path(tempdir(), "lexsync_demo")
dir.create(out, showWarnings = FALSE)
files <- export_experiments(stim, design, schema, out)
basename(unlist(files))
```

The PsychoPy script binds each onset trigger to the stimulus flip with
`win.callOnFlip`; the OpenSesame `.osexp` sends the marker from the item that
immediately follows the word. Neither the matching nor the script generation
requires PsychoPy, OpenSesame or any hardware to be installed.
